---
title: "6 - JSDM Occurrence (probit) model results (best WAIC)"
output: html_document
date: "2026-04-14"
editor_options:
  chunk_output_type: inline
---

This script analyses the second best probit (occurrence) JSDM selected by WAIC 32 additive covariate combinations fitted in `5_jsdm_genus_models.R` then by cross-validation from the 
`6.5_cross_validation_best_models.Rmd`


# 1 -- Packages & themes

```{r, include=FALSE}
library(tidyverse)
library(raster)
library(pals)
library(viridis)
library(Hmsc)
library(ape)
library(ggtree)
library(nlme)
library(ggspatial)
library(patchwork)
library(corrplot)
library(sf)
library(magrittr)

source("code/ggheatmap.custom.R")

main_theme <- theme_minimal() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(colour = "black", size = 13, face = "italic",
                                   angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(colour = "black", size = 13, face = "italic"),
        axis.title = element_text(colour = "black", size = 12),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        legend.title = element_text(colour = "black", size = 12, hjust = 0.5),
        legend.text = element_text(colour = "black", size = 13))

main_theme_final <- theme_bw() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(colour = "black", size = 7, face = "italic",
                                   vjust = 1, hjust = 1),
        axis.text.y = element_text(colour = "black", size = 7, face = "italic"),
        axis.title = element_text(colour = "black", size = 9),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_text(colour = "black", size = 9, hjust = 0.5),
        legend.key.size = unit(0.3, "cm"),
        legend.text = element_text(colour = "black", size = 7, face = "italic"),
        strip.background = element_blank(),
        strip.text = element_text(margin = margin(0.05, 0, 0.05, 0, "cm"), size = 7),
        title = element_text(colour = "black", size = 9))

color_palette <- c(
  "Biomass quantity index" = "#E69F00",
  "Precipitation"          = "#56B4E9",
  "T\u00b0 max"            = "#009E73",
  "T\u00b0 delta"          = "#F0E442",
  "Wind"                   = "#0072B2",
  "Variance not explained" = "#999999"
)
```

# 2 -- Load data & models

```{r load-data}
covariate_standa_df <- read.table("outputs/data_for_models/covariate_standa_df.txt",
                                  stringsAsFactors = TRUE)
Y_genus   <- read.table("outputs/data_for_models/Y_genus.txt")
genus_tree <- ape::read.tree("outputs/data_for_models/genus_tree.tree")
```

# 3 -- Rename genus

```{r rename-genus}
genus_rename <- genus_tree$tip.label %>%
  str_remove("g__") %>%
  str_replace_all("[a-z]__", "Unk. ") %>%
  str_replace("_(\\d+)$", " (\\1 OTU)") %>%
  str_replace("_", " ")

genus_tree$tip.label <- genus_rename

Y_genus <- Y_genus %>%
  rename_with(. %>%
                str_remove("g__") %>%
                str_replace_all("[a-z]__", "Unk. ") %>%
                str_replace("_(\\d+)$", " (\\1 OTU)") %>%
                str_replace("_", " "))

Y_genus <- Y_genus[, genus_tree$tip.label]
```

```{r load-models}
load(file = "outputs/models/list_mods_bin_phylo_autocorr_genus.RData")
```

# 4 -- Select best model by WAIC

```{r compute-waic}
if(file.exists("outputs/models/WAIC_bin_mod.txt")){
  WAIC_bin_mod <- read.table("outputs/models/WAIC_bin_mod.txt")
}else{
  model_names_bin <- names(list_mods_bin)
  WAIC_vector     <- numeric(length(list_mods_bin))
  
  for (m in seq_along(list_mods_bin)) {
    WAIC_vector[m] <- computeWAIC(list_mods_bin[[m]]$m_converg)
  }
  
  WAIC_bin_mod <- data.frame(
    WAIC_vecor   = WAIC_vector,
    model_names = model_names_bin
  )

  write.table(WAIC_bin_mod, file = "outputs/models/WAIC_bin_mod.txt")
}
```

```{r select-best-model}
WAIC_bin_mod %>%
  arrange(WAIC_vecor_bin) %>%
  knitr::kable(caption = "All occurrence models ranked by WAIC (lower is better)")

best_model_name <- WAIC_bin_mod %>%
  slice_min(WAIC_vecor_bin, n = 2) %>%
  pull(model_names_bin)

cat("Best model selected:", best_model_name, "\n")
cat("WAIC:", WAIC_bin_mod %>%
      slice_min(WAIC_vecor_bin, n = 2) %>%
      pull(WAIC_vecor_bin), "\n")
```

```{r extract-best-model}
m_best <- list_mods_bin[[best_model_name[2]]]$m_converg
```

# 5 -- Convergence diagnostics

```{r convergence-diagnostics}
postBeta <- getPostEstimate(m_best, parName = "Beta")
mpost    <- convertToCodaObject(m_best)

# Gelman-Rubin PSRF
gelman_diag <- gelman.diag(mpost$Beta, multivariate = FALSE)
hist(gelman_diag$psrf[, 1], breaks = 50,
     main = paste("Gelman-Rubin PSRF -", best_model_name),
     xlab = "PSRF (should be < 1.1)")

# Effective sample size
ess <- effectiveSize(mpost$Beta)
hist(ess, breaks = 50,
     main = paste("Effective sample size -", best_model_name),
     xlab = "Effective sample size")

# Phylogenetic signal parameter (Rho)
cat("\n--- Rho convergence ---\n")
gelman.diag(mpost$Rho, multivariate = FALSE)
summary(mpost$Rho)
```

# 6 -- Beta heatmap on phylogenetic tree

```{r beta-heatmap, echo=FALSE, fig.height=8, fig.width=9}
beta_results <- postBeta$mean
rownames(beta_results) <- colnames(m_best$X)
colnames(beta_results) <- str_replace(colnames(beta_results), "Unknown ", "Unk. ")
colnames(beta_results) <- str_replace(colnames(beta_results), "_", " ")

# Mask non-significant effects (95% posterior support)
beta_results[!(postBeta$support > 0.95 | postBeta$support < 0.05)] <- NA

# Reorder to match phylogenetic tree
beta_results <- t(beta_results)[genus_tree$tip.label, colnames(m_best$X)]

# Rename covariates for display
covariate_labels <- c("(Intercept)" = "Intercept",
                      "ndvi"        = "NDVI",
                      "precip"      = "Precip.",
                      "tmax"        = "T\u00b0 max",
                      "tdelta"      = "T\u00b0 delta",
                      "wind"        = "Wind spd")

existing_cols <- colnames(beta_results)
new_col_names <- covariate_labels[existing_cols]
new_col_names[is.na(new_col_names)] <- existing_cols[is.na(new_col_names)]
colnames(beta_results) <- new_col_names

# Phylo-heatmap
p <- ggtree(genus_tree, layout = "roundrect") +
  geom_tiplab(align = TRUE, size = 3) +
  theme_tree()

ggheatmap.custom(p = p, data = beta_results, offset = 1, width = 1,
                 font.size = 5, colnames_angle = -60, hjust = 0) +
  scale_fill_gradient2(low = "#366785", mid = "white", high = "#c85200",
                       na.value = "grey90",
                       limits = c(min(beta_results, na.rm = TRUE),
                                  max(beta_results, na.rm = TRUE))) +
  scale_y_continuous(expand = c(0, 5)) +
  labs(fill = "Parameters\nvalues") +
  main_theme_final +
  theme_void() -> effect_plot_bin

ggsave(plot = effect_plot_bin,
       filename = "figures/genus_bin_autocorr/effect_plot_bin_best_waic.pdf",
       units = "cm", height = 18, width = 20)

effect_plot_bin
```

# 7 -- Phylogenetic signal (Pagel's lambda)

```{r pagel-lambda, include=FALSE}
compute.pagel.lambda <- function(model) {
  postBeta  <- getPostEstimate(model, parName = "Beta")
  phyloTree <- model$phyloTree
  nc <- dim(postBeta$mean)[1]

  results <- data.frame(
    lambda    = numeric(nc),
    p_value   = numeric(nc),
    covariate = character(nc),
    stringsAsFactors = FALSE
  )

  control <- glsControl(opt = "optim", msVerbose = FALSE)

  for (j in 1:nc) {
    y  <- postBeta$mean[j, ]
    m1 <- gls(y ~ 1)
    m2 <- tryCatch(
      gls(y ~ 1,
          correlation = corPagel(0.5, phyloTree, fixed = FALSE,
                                 form = ~names(y)),
          control = control),
      error = function(e) NULL
    )

    if (!is.null(m2)) {
      z1 <- summary(m2)
      z2 <- anova(m1, m2)
      results$lambda[j]  <- z1$modelStruct[1]$corStruct
      results$p_value[j] <- z2$`p-value`[2]
    } else {
      results$lambda[j]  <- NA
      results$p_value[j] <- NA
    }
    results$covariate[j] <- colnames(model$X)[j]
  }
  results
}

lambda_df <- compute.pagel.lambda(m_best)
```

```{r pagel-lambda-table, echo=FALSE}
lambda_df %>%
  knitr::kable(caption = "Pagel's lambda for each covariate")
```

```{r pagel-lambda-plot, echo=FALSE, fig.height=6, fig.width=6}
lambda_df %>%
  mutate(covariate = case_when(
    covariate == "(Intercept)" ~ "Intercept",
    covariate == "ndvi"        ~ "NDVI",
    covariate == "precip"      ~ "Precip.",
    covariate == "tmax"        ~ "T\u00b0 max",
    covariate == "tdelta"      ~ "T\u00b0 delta",
    covariate == "wind"        ~ "Wind spd",
    .default = covariate
  )) %>%
  ggplot() +
  geom_point(aes(lambda, covariate, alpha = p_value <= 0.05),
             cex = 4, shape = 21, fill = "#0072B2") +
  geom_vline(xintercept = 1, linetype = 2) +
  labs(x = "Phylogenetic structure of responses",
       y = "Covariates", alpha = "p <= 0.05") +
  theme_classic() +
  theme(aspect.ratio = 1,
        axis.text.x  = element_text(colour = "black", size = 11, face = "italic",
                                    angle = 18, vjust = 1, hjust = 1),
        axis.text.y  = element_text(colour = "black", size = 11, face = "italic"),
        axis.title   = element_text(colour = "black", size = 14),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        legend.title = element_text(colour = "black", size = 14, hjust = 0.5),
        legend.text  = element_text(colour = "black", size = 11)) -> phylo_signal_bin

ggsave(plot = phylo_signal_bin,
       filename = "figures/genus_bin_autocorr/phylogenetic_signal_bin_best_waic.pdf",
       units = "cm", height = 16, width = 16)

phylo_signal_bin
```

# 8 -- Variance partitioning

```{r variance-partitioning, echo=FALSE, warning=FALSE, fig.height=8, fig.width=8}
VP <- computeVariancePartitioning(m_best, na.ignore = TRUE)

plot_data_vp <- VP$vals %>%
  as.data.frame() %>%
  rownames_to_column("var") %>%
  mutate(var = case_when(
    var == "ndvi"   ~ "Biomass quantity index",
    var == "precip" ~ "Precipitation",
    var == "tmax"   ~ "T\u00b0 max",
    var == "tdelta" ~ "T\u00b0 delta",
    var == "wind"   ~ "Wind",
    .default = var
  ),
  var = factor(var, levels = rev(c("Biomass quantity index", "Precipitation",
                                   "T\u00b0 max", "T\u00b0 delta", "Wind",
                                   "Random: sample", "Random: sample_id")))) %>%
  pivot_longer(-var) %>%
  mutate(name = str_replace(name, "_", " "),
         name = str_replace(name, "Unknown ", "Unk. "))

ggplot(plot_data_vp, aes(x = value, y = name, fill = var)) +
  geom_col(color = "white", linewidth = 0.2) +
  geom_label(data = subset(plot_data_vp, var == levels(plot_data_vp$var)[1]),
             aes(label = name),
             x = 0.01, hjust = 0, size = 2.5, fontface = "italic",
             fill = "white", alpha = 0.5, label.size = NA,
             label.padding = unit(0.15, "lines")) +
  scale_fill_manual(values = color_palette, na.value = "grey70") +
  scale_x_continuous(expand = c(0, 0), labels = scales::percent_format()) +
  labs(x = "Proportion of Variance Explained",
       y = NULL,
       fill = "Environmental Covariates",
       title = paste("Variance partitioning -", best_model_name)) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x  = element_text(color = "black", size = 10),
    axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10)),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text  = element_text(size = 10),
    plot.margin  = margin(10, 20, 10, 10),
    aspect.ratio = 1
  ) -> vp_bin

ggsave(plot = vp_bin,
       filename = "figures/genus_bin_autocorr/variance_partitioning_bin_best_waic.pdf",
       units = "cm", height = 16, width = 16)

vp_bin
```

# 9 -- Residual associations (Omega)

Correlation between each genus when all the environmental variables have been
accounted for. These figures show unexplained correlations caused by omitted
biotic or abiotic variables.

```{r residual-associations, echo=FALSE, fig.height=8, fig.width=8}
OmegaCor     <- computeAssociations(m_best)
supportLevel <- 0.95

for (i in seq_along(OmegaCor)) {
  toPlot <- ((OmegaCor[[i]]$support > supportLevel) +
               (OmegaCor[[i]]$support < (1 - supportLevel)) > 0) * OmegaCor[[i]]$mean

  corrplot(toPlot, method = "color",
           col = colorRampPalette(c("blue", "white", "red"))(200),
           tl.cex = 0.6, tl.col = "black",
           title = paste("Residual associations -", m_best$rLNames[i]),
           mar = c(0, 0, 1, 0), order = "hclust")
}
```

# 10 -- Taxonomy reference

```{r taxonomy, message=FALSE}
data_18S <- "data/18S_AMf_Eukaryome/Robust_MiSeq_18S_AMf_20241120_125_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.table2"
amf_taxonomic_levels_eucaryome <- c("kingdom", "phylum", "class", "order",
                                    "family", "genus", "species")

taxomony_AMF_genuslevel <- data_18S %>%
  read_tsv() %>%
  dplyr::select(taxonomy) %>%
  mutate(taxonomy = str_replace_all(taxonomy, ":", "_"),
         taxonomy = str_replace_all(taxonomy, "[|]", "; ")) %>%
  filter(str_detect(taxonomy, "p__Glomeromycota")) %>%
  mutate(taxonomy = str_replace_all(taxonomy,
           "(p__|c__|o__|f__|g__|s__)unclassified", "*")) %>%
  separate_wider_delim(taxonomy, names = amf_taxonomic_levels_eucaryome,
                       delim = "; ", cols_remove = FALSE) %>%
  unite(taxonomy, kingdom, phylum, class, order, family, genus,
        sep = "; ", remove = FALSE) %>%
  group_by(taxonomy) %>%
  mutate(
    genus = paste(str_extract(
      taxonomy,
      "(g_[^;*]+$)|([a-z]_[^;*]+)(?=(;\\s*\\*|\\s*\\*|$))"),
      n(), sep = "_")
  ) %>%
  group_by(genus) %>%
  summarise(kingdom = first(kingdom),
            phylum  = first(phylum),
            class   = first(class),
            order   = first(order),
            family  = first(family)) %>%
  mutate(genus = genus %>%
           str_remove("g__") %>%
           str_replace_all("[a-z]__", "Unk. ") %>%
           str_replace("_(\\d+)$", " (\\1 OTU)"))
```

# 11 -- Spatial prediction maps (limited gradient)

## 11.1 Prepare raster covariates

```{r prepare-rasters, echo=FALSE}
ndvi_uganda              <- raster("outputs/raster/ndvi_uganda_lower_res.tif")
precip_uganda_resampled  <- raster("outputs/raster/precip_2023_lower_res.tif")
tmax_uganda_resampled    <- raster("outputs/raster/tmax_2023_lower_res.tif")
tmin_uganda_resampled    <- raster("outputs/raster/tmin_2023_lower_res.tif")
delta_t_uganda           <- tmax_uganda_resampled - tmin_uganda_resampled
wind_2023                <- raster("outputs/raster/wind_2023.tif")

ndvi_raster_df <- rasterToPoints(ndvi_uganda) %>%
  as.data.frame() %>% rename(ndvi = "ndvi_uganda_lower_res")

precip_raster_df <- rasterToPoints(precip_uganda_resampled) %>%
  as.data.frame() %>% rename(precip = "precip_2023_lower_res")

tmax_raster_df <- rasterToPoints(tmax_uganda_resampled) %>%
  as.data.frame() %>% rename(tmax = "tmax_2023_lower_res")

tdelta_raster_df <- rasterToPoints(delta_t_uganda) %>%
  as.data.frame() %>% rename(tdelta = "layer")

wind_raster_df <- rasterToPoints(wind_2023) %>%
  as.data.frame() %>% rename(wind = "wind_2023")

data_spatial_df <- ndvi_raster_df %>%
  left_join(precip_raster_df, by = c("x", "y")) %>%
  left_join(tmax_raster_df,   by = c("x", "y")) %>%
  left_join(tdelta_raster_df, by = c("x", "y")) %>%
  left_join(wind_raster_df,   by = c("x", "y")) %>%
  na.omit() %>%
  mutate(across(c(ndvi, precip, tmax, tdelta, wind),
                ~ vegan::decostand(.x, method = "standardize")[, 1]))
```

```{r environmental-maps, echo=FALSE, fig.height=8, fig.width=14}
data_spatial_df %>%
  pivot_longer(-c(x, y)) %>%
  mutate(name = case_when(
    name == "ndvi"   ~ "Biomass quantity index",
    name == "precip" ~ "Precipitation",
    name == "tmax"   ~ "T\u00b0 max",
    name == "tdelta" ~ "T\u00b0 delta",
    name == "wind"   ~ "Wind"),
    name = factor(name, levels = c("Biomass quantity index", "Precipitation",
                                   "T\u00b0 max", "T\u00b0 delta", "Wind"))) %>%
  ggplot() +
  facet_wrap(~name) +
  geom_raster(aes(x, y, fill = value)) +
  scale_fill_viridis(option = "viridis") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Environmental\nstandardized values") +
  main_theme_final -> environmental_maps

ggsave(plot = environmental_maps,
       filename = "figures/genus_bin_autocorr/environmental_maps_bin.pdf",
       units = "cm", width = 33, height = 20)

environmental_maps
```

## 11.2 Limited environmental gradient

```{r limited-gradient}
ug_border <- read_sf("data/ug_shp/ug.shp")

data_spatial_limited_df <- data_spatial_df %>%
  filter(
    max(covariate_standa_df$ndvi)   > ndvi   & min(covariate_standa_df$ndvi)   < ndvi,
    max(covariate_standa_df$precip) > precip & min(covariate_standa_df$precip) < precip,
    max(covariate_standa_df$tmax)   > tmax   & min(covariate_standa_df$tmax)   < tmax,
    max(covariate_standa_df$tdelta) > tdelta & min(covariate_standa_df$tdelta) < tdelta,
    max(covariate_standa_df$wind)   > wind   & min(covariate_standa_df$wind)   < wind
  )
```

## 11.3 Occurrence prediction

```{r occurrence-prediction, echo=FALSE}
xy.grid_limited <- data_spatial_limited_df %>%
  dplyr::select(x, y) %>%
  as.matrix()

XData.grid_limited <- data_spatial_limited_df %>%
  dplyr::select(-c(x, y))

# Only keep covariates present in the best model
model_covariates <- setdiff(colnames(m_best$X), "(Intercept)")
XData.grid_limited <- XData.grid_limited[, model_covariates, drop = FALSE]

save_path <- "outputs/models/occurrence_prediction_bin_best_waic.RData"

if (file.exists(save_path)) {
  load(save_path)
} else {
  Gradient_limited <- prepareGradient(
    m_best,
    XDataNew = XData.grid_limited,
    sDataNew = list(sample = xy.grid_limited))

  predY_limited <- stats::predict(m_best, Gradient = Gradient_limited,
                                  predictEtaMean = TRUE, expected = TRUE)

  EpredY_map_limited <- predY_limited[[1]]
  if (length(predY_limited) > 1) {
    for (i in 2:length(predY_limited)) {
      EpredY_map_limited <- EpredY_map_limited + predY_limited[[i]]
      if (i %% 1000 == 0) print(i)
    }
    EpredY_map_limited <- EpredY_map_limited / length(predY_limited)
  }
  save(EpredY_map_limited, file = save_path)
}
```

```{r compute-thresholds}
# Load cross-validation predictions for the second best binary model
load("outputs/models/cv_bin_second_best_waic.RData")

compute.ROC <- function(observed, predicted, pred_name) {
  df_roc_AUC <- data.frame(
    threshold = sort(unique(predicted)),
    pred_name = pred_name,
    precision = NA, TPR = NA, FPR = NA, F1_score = NA
  )
  for (n in 1:nrow(df_roc_AUC)) {
    th <- df_roc_AUC[n, 1]
    temp_pred <- predicted > th
    TP <- sum(observed & temp_pred)
    TN <- sum(!(observed | temp_pred))
    FP <- sum(!observed & temp_pred)
    FN <- sum(observed & !temp_pred)
    precision <- TP / (TP + FP)
    TPR <- TP / (TP + FN)
    FPR <- FP / (FP + TN)
    df_roc_AUC[n, "precision"] <- precision
    df_roc_AUC[n, "TPR"]       <- TPR
    df_roc_AUC[n, "FPR"]       <- FPR
    df_roc_AUC[n, "F1_score"]  <- 2 * (precision * TPR) / (precision + TPR)
  }
  return(df_roc_AUC)
}

genus_with_variation <- colnames(m_best$Y)[
  colSums(m_best$Y) > 0 & colSums(m_best$Y) < nrow(m_best$Y)
]

1:ncol(m_best$Y) %>%
  # purrr::keep(~ colnames(m_best$Y)[.x] %in% genus_with_variation) %>%
  purrr::map(~ compute.ROC(
    observed  = m_best$Y[, .x],
    predicted = mPredY_bin2[, .x],
    pred_name = colnames(m_best$Y)[.x]
  )) %>%
  bind_rows() -> ROC_df

# Best balanced threshold per genus (max Youden's J = TPR - FPR)
ROC_df %>%
  mutate(balanced_score = TPR - FPR) %>%
  group_by(pred_name) %>%
  filter(balanced_score == max(balanced_score, na.rm = F)) %>%
  filter(TPR == max(TPR, na.rm = TRUE)) %>%
  dplyr::select(pred_name, threshold, balanced_score, TPR, FPR) %>%
  rename(TPR_b = "TPR", FPR_b = "FPR") %>%
  distinct() %>%
# Arrange the threshold manually for 4 genus that are always present.
  rbind(.,data.frame(pred_name =  c("Acaulospora (88 OTU)", 
                                                "Unk. Glomeraceae (120 OTU)"), threshold = 0, balanced_score =NA)) %>%
  mutate(threshold = case_when(pred_name %in% c("Nanoglomus (128 OTU)",
                                              "Sclerocystis (52 OTU)") ~ 0,
                               .default = threshold),
         TPR_b = case_when(pred_name %in% c("Nanoglomus (128 OTU)",
                                              "Sclerocystis (52 OTU)") ~ NA,
                           .default = TPR_b),
         FPR_b = case_when(pred_name %in% c("Nanoglomus (128 OTU)",
                                              "Sclerocystis (52 OTU)") ~ NA,
                           .default = FPR_b),
                 ) -> balanced_score_df
```

```{r occurrence-map-data}
colnames(EpredY_map_limited) <- colnames(EpredY_map_limited) %>%
  str_replace("Unknown ", "Unk. ") %>%
  str_replace("_", " ")

balanced_score_df <- balanced_score_df %>%
  mutate(pred_name = str_replace(pred_name, "_", " "),
         pred_name = str_replace(pred_name, "Unknown ", "Unk. "))

balanced_score_df$pred_name %>% length()
colnames(EpredY_map_limited) %in% balanced_score_df$pred_name
EpredY_map_limited <- EpredY_map_limited[, balanced_score_df$pred_name]

# Threshold probabilities to presence/absence
EpredY_occurrence_map <- sweep(EpredY_map_limited, 2,
                               balanced_score_df$threshold,
                               FUN = function(x, th) x > th)
EpredY_occurrence_map <- +EpredY_occurrence_map

data_spatial_limited_df %>%
  bind_cols(as.data.frame(EpredY_occurrence_map)) %>%
  mutate(pixel_ID = as.factor(1:n())) %>%
  pivot_longer(-c(pixel_ID, x, y, ndvi, precip, tmax, tdelta, wind),
               values_to = "occurrence", names_to = "genus") %>%
  mutate(occurrence = as.factor(occurrence)) %>%
  left_join(taxomony_AMF_genuslevel, by = "genus") -> EpredY_occurrence_df_limited
```

## 11.4 All-genus occurrence map

```{r occurrence-map-all, echo=FALSE, fig.height=12, fig.width=12}
EpredY_occurrence_df_limited %>%
  ggplot() +
  facet_wrap(~genus) +
  geom_raster(aes(x, y, fill = occurrence)) +
  scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0.05, end = 0.6) +
  labs(x = "Longitude", y = "Latitude", fill = "Occurrence",
       title = paste("Predicted occurrence -", best_model_name)) +
  main_theme_final -> map_occurrence_all

map_occurrence_all
```

## 11.5 Per-family occurrence maps

### Glomeraceae

```{r occurrence-glomeraceae, echo=FALSE, fig.height=10, fig.width=10}
EpredY_occurrence_df_limited %>%
  filter(family == "f__Glomeraceae") %>%
  ggplot() +
  geom_sf(data = ug_border, alpha = 0.4) +
  facet_wrap(~genus) +
  geom_tile(aes(x, y, fill = occurrence)) +
  scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0.05, end = 0.6) +
  labs(x = "Longitude", y = "Latitude", fill = "Occurrence",
       title = "Glomeraceae") +
  main_theme_final -> occurrence_Glomeraceae_map

ggsave(plot = occurrence_Glomeraceae_map,
       filename = "figures/genus_bin_autocorr/occurrence_Glomeraceae_bin_best_waic.pdf",
       units = "cm", width = 48, height = 48)

occurrence_Glomeraceae_map
```

### Gigasporaceae

```{r occurrence-gigasporaceae, echo=FALSE, fig.height=5, fig.width=5}
EpredY_occurrence_df_limited %>%
  filter(family == "f__Gigasporaceae") %>%
  ggplot() +
  geom_sf(data = ug_border, alpha = 0.4) +
  facet_wrap(~genus) +
  geom_tile(aes(x, y, fill = occurrence)) +
  scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0.05, end = 0.6) +
  labs(x = "Longitude", y = "Latitude", fill = "Occurrence",
       title = "Gigasporaceae") +
  main_theme_final -> occurrence_Gigasporaceae_map

ggsave(plot = occurrence_Gigasporaceae_map,
       filename = "figures/genus_bin_autocorr/occurrence_Gigasporaceae_bin_best_waic.pdf",
       units = "cm", width = 12, height = 12)

occurrence_Gigasporaceae_map
```

### Acaulosporaceae

```{r occurrence-acaulosporaceae, echo=FALSE, fig.height=5, fig.width=5}
EpredY_occurrence_df_limited %>%
  filter(family == "f__Acaulosporaceae") %>%
  ggplot() +
  geom_sf(data = ug_border, alpha = 0.4) +
  facet_wrap(~genus) +
  geom_tile(aes(x, y, fill = occurrence)) +
  scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0.05, end = 0.6) +
  labs(x = "Longitude", y = "Latitude", fill = "Occurrence",
       title = "Acaulosporaceae") +
  main_theme_final -> occurrence_Acaulosporaceae_map

ggsave(plot = occurrence_Acaulosporaceae_map,
       filename = "figures/genus_bin_autocorr/occurrence_Acaulosporaceae_bin_best_waic.pdf",
       units = "cm", width = 12, height = 12)

occurrence_Acaulosporaceae_map
```

### Diversisporaceae

```{r occurrence-diversisporaceae, echo=FALSE, fig.height=12, fig.width=5}
EpredY_occurrence_df_limited %>%
  filter(family == "f__Diversisporaceae") %>%
  group_split(genus) %>%
  purrr::map(~ {
    ggplot() +
      facet_wrap(~genus) +
      geom_sf(data = ug_border, alpha = 0.4) +
      geom_tile(data = .x, aes(x, y, fill = occurrence)) +
      scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0.05, end = 0.6) +
      labs(x = "Longitude", y = "Latitude", fill = "Occurrence") +
      main_theme_final
  }) %>%
  wrap_plots(nrow = 3) -> occurrence_Diversisporaceae_map

ggsave(plot = occurrence_Diversisporaceae_map,
       filename = "figures/genus_bin_autocorr/occurrence_Diversisporaceae_bin_best_waic.pdf",
       units = "cm", width = 12, height = 36)

occurrence_Diversisporaceae_map
```

# 12 -- Richness map (limited gradient)

```{r richness-map, echo=FALSE, fig.height=6, fig.width=6}
map_ug <- rnaturalearth::ne_states("Uganda", returnclass = "sf")

EpredY_occurrence_df_limited %>%
  filter(!str_detect(genus, "Unk.")) %>%
  group_by(pixel_ID) %>%
  summarise(richness = sum(as.numeric(as.character(occurrence))),
            x = first(x),
            y = first(y)) %>%
  ggplot() +
  geom_sf(data = map_ug, alpha = 0.4) +
  geom_tile(aes(x, y, fill = richness), alpha = 0.9) +
  annotation_north_arrow(
    location = "tr", which_north = "true",
    height = unit(1, "cm"), width = unit(1, "cm"),
    style = north_arrow_minimal(
      line_col = "black", fill = "black",
      text_col = "black", line_width = 2, text_size = 10)) +
  scale_fill_viridis(option = "magma") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Predicted genus\nrichness") +
  main_theme_final -> map_richness_bin

ggsave(plot = map_richness_bin,
       filename = "figures/genus_bin_autocorr/map_richness_bin_best_waic.pdf",
       units = "cm", width = 12, height = 12)

map_richness_bin
```

# 13 -- Session info

```{r session-info}
cat("Best model:", best_model_name, "\n")
cat("Model covariates:", paste(colnames(m_best$X), collapse = ", "), "\n")
cat("Distribution: Probit (occurrence)\n")
cat("Number of species:", ncol(m_best$Y), "\n")
cat("Number of sites:", nrow(m_best$Y), "\n")
sessionInfo()
```
