---
title: "Extract data from stac API"
output: html_document
date: "2025-05-13"
---
```{r}
library(rstac)
library(magrittr)
library(terra)
library(raster)
library(stars)
```

# NDVI raster
## Prepartation before loading raster
A STAC URl where data catalog is located
```{r}
s_obj <- stac("https://planetarycomputer.microsoft.com/api/stac/v1/")


#Choose the location of the data (WGS) and the time
uganda_bbox <- c(29.5, -1.5, 35.0, 4.5) #
start_date <- as.Date("2022-06-01")
end_date   <- as.Date("2023-05-31")
# start_date <-as.Date("2023-07-01")
# end_date <-as.Date("2023-07-10")
```

Choose collection and load all information of all available dataset for the choosen location and time period

```{r}
it_obj <- s_obj %>%
  stac_search(
    collections = "sentinel-3-synergy-v10-l2-netcdf",
    bbox = uganda_bbox,
    datetime = paste0(as.character(start_date), "/", as.character(end_date))
  ) %>%
  get_request()
```

## Load raster in a temporary file
```{r}
signed_items <- it_obj %>% items_sign(sign_planetary_computer())
tmp_file_list = c()
for(n in 1:length(signed_items$features)){
  ndvi_url <- signed_items$features[[n]]$assets$ndvi$href
  tmp_file_list <-c(tmp_file_list, tempfile(fileext = ".nc")) 
  download.file(ndvi_url, destfile = tmp_file_list[n], mode = "wb")
}

rasters <- stack(tmp_file_list)
ndvi_mean <- calc(rasters, mean, na.rm = TRUE)
plot(ndvi_mean)
```


## Crop an resize raster 

```{r}
# Define Uganda extent as an sf object
uganda_bbox_sf <- sf::st_as_sfc(sf::st_bbox(c(
  xmin = 29.5, ymin = -1.5, xmax = 35.0, ymax = 4.5
), crs = st_crs(ndvi_mean)))

# Crop the raster
raster::crop()
ndvi_mean_uganda <- ndvi_mean %>%
  crop(extent(29.5, 35.0, -1.5, 4.5)) %>%
  as( "Raster")
 

plot(ndvi_mean_uganda)
```


## Save raster  

```{r}
writeRaster(ndvi_mean_uganda,"data/raster/ndvi_mean_uganda.tif",options=c('TFW=YES'))
```


# Land cover raster
## Prepartation before loading raster
A STAC URl where data catalog is located
```{r}
s_obj <- stac("https://planetarycomputer.microsoft.com/api/stac/v1/")


#Choose the location of the data (WGS) and the time
uganda_bbox <- c(29.5, -1.5, 35.0, 4.5) #

start_date <-as.Date("2023-07-01")
end_date <-as.Date("2023-07-31")
```

Choose collection and load all information of all available dataset for the choosen location and time period

```{r}
it_obj <- s_obj %>%
  stac_search(
    collections = "sentinel-3-slstr-lst-l2-netcdf",
    bbox = uganda_bbox,
    datetime = paste0(as.character(start_date), "/", as.character(end_date))
  ) %>%
  get_request()
```

## Load raster in a temporary file
```{r}
signed_items <- it_obj %>% items_sign(sign_planetary_computer())

surface_t_url1 <- signed_items$features[[1]]$assets$`lst-in`$href
surface_t_url2 <- signed_items$features[[2]]$assets$`lst-in`$href
tmp_file2 <- tempfile(fileext = ".nc")
tmp_file3 <- tempfile(fileext = ".nc")
download.file(surface_t_url1, destfile = tmp_file2, mode = "wb")
download.file(surface_t_url2, destfile = tmp_file3, mode = "wb")
```


```{r}
library(stars)
surface_t_raster1 <- read_stars(tmp_file2)
surface_t_raster2 <- read_stars(tmp_file3)

lst1 <- surface_t_raster1["LST"] %>%
  as("Raster")
lst2 <- surface_t_raster2["LST"]%>%
  as("Raster")
s <- merge(lst2, lst1)
par(mfrow = c(1,2))
plot(s)
plot(lst1)
```

```{r}
uganda_bbox_sf <- sf::st_as_sfc(sf::st_bbox(c(
  xmin = 29.5, ymin = -1.5, xmax = 35.0, ymax = 4.5
), crs = st_crs(rast_UTM_36N)))
temp <- rast_UTM_36N %>%
  raster() %>%
  ratify()
# Crop the raster
levels(temp)
rast_UTM_36N_cover <- temp %>%
  raster::aggregate(fact=100, fun = modal,  na.rm=TRUE)
plot(rast_UTM_36N_cover)
```

```{r}
library(mapview)
library(leaflet)
library(leafem)
library(rasterVis)

leaflet() %>%
  addTiles() %>% 
  addRasterImage(rast_UTM_36N_cover, opacity = 0.7)
```

## Crop an resize raster 

```{r}
# Define Uganda extent as an sf object
uganda_bbox_sf <- sf::st_as_sfc(sf::st_bbox(c(
  xmin = 29.5, ymin = -1.5, xmax = 35.0, ymax = 4.5
), crs = st_crs(ndvi_raster1)))

# Crop the raster
ndvi_uganda <- ndvi_raster1 %>%
  st_crop(uganda_bbox_sf) %>%
  as( "Raster")

plot(ndvi_uganda)
```

## Save raster  

```{r}
writeRaster(ndvi_uganda,"data/raster/ndvi_uganda.tif",options=c('TFW=YES'))
```