---
title: "jsdm_family"
output: html_document
date: "2025-05-27"
editor_options: 
  chunk_output_type: inline
---
```{r}
library(tidyverse)
library(raster)
library(mapview)
library(leaflet)
library(leafem)
library(rasterVis)
library(readxl)
library(sf)
library(pals)

library(Hmsc)
library(ape)

```

# Covariate

```{r echo=FALSE}
metadata <- read_xlsx(path = "data/Metadata_Doreen_070525.xlsx")

utm_coords <- metadata %>%
  mutate(
    zone = as.numeric(substr(easting, 1, 2)),
    band = substr(easting, 3, 3),
    easting = as.numeric(sub(".* ", "", easting)),
    northing = as.numeric(northing),

    hemisphere = ifelse(band == "N", "north", "south") # M = South Hemisphere
  )

# Function to convert a single row to lat/lon
convert_to_wgs84 <- function(zone, hemisphere, easting, northing,Sample_code) {
  epsg <- if (hemisphere == "north") {
    32600 + zone  # Northern hemisphere UTM EPSG codes
  } else {
    32700 + zone  # Southern hemisphere UTM EPSG codes
  }

  point <- st_sfc(st_point(c(easting, northing)), crs = epsg)
  point_wgs84 <- st_transform(point, crs = 4326)
  coords <- st_coordinates(point_wgs84)
  tibble( lon = coords[1], lat = coords[2],Sample_code = Sample_code)
}
utm_coords %>%
  mutate(zone = case_when(zone==32~ 36,
                             .default = zone),
          northing = case_when(Sample_code == "N17" ~ 86562,
                             .default = northing)) -> utm_coords
# Apply to each row
wgs84_coords <- pmap_dfr(utm_coords[, c("zone", "hemisphere", "easting", "northing","Sample_code")], convert_to_wgs84) %>%
  left_join(metadata, by = "Sample_code") %>%
  st_as_sf(coords = c("lon", "lat"),crs = 4326) %>%
  mutate(
     lon= st_coordinates(.)[,1],
     lat= st_coordinates(.)[,2]
  )


leaflet() %>% 
  addTiles() %>%
  addMarkers(data=wgs84_coords,~lon, ~lat, popup = ~paste(round(lat,2), round(lon,2)) , label = ~Sample_code)%>%
  addCircleMarkers(data = wgs84_coords, lng = ~lon,
                  lat = ~lat, radius = 8, color = "blue", fillOpacity = 0.8)
```


```{r echo=FALSE}
land_degradation_raster <- raster("outputs/raster/land_degradation_lower_res.tif")
land_degradation_type <- levels(land_degradation_raster)[[1]]

land_degradation_df <- raster::extract(land_degradation_raster,  wgs84_coords)%>%
  tibble(ID = .,
  Sample_code  = wgs84_coords$Sample_code)%>%
  left_join(land_degradation_type,  by = "ID")

myPal1 <- rev(RColorBrewer::brewer.pal('PRGn', n=5))


leaflet() %>%
  addTiles() %>% 
  addRasterImage(land_degradation_raster, opacity = 0.7,
                 color = myPal1) %>%
   addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal1, domain = land_degradation_type$land_degra),
    values = land_degradation_type$land_degra,
    title = "Land Degradation"
  )
```


```{r echo=FALSE}
land_cover_raster <- raster("outputs/raster/land_cover_lower_res.tif")

land_cover_type <- levels(land_cover_raster)[[1]] %>%
  dplyr::select(ID,Land_Cover)


land_cover_df <- raster::extract(land_cover_raster, wgs84_coords)%>%
 tibble(ID = .,
    Sample_code  = wgs84_coords$Sample_code) %>%
    left_join(land_cover_type, by = "ID")

myPal2 <- pals::stepped2(n=18)

leaflet() %>%
  addTiles() %>% 
  addRasterImage(land_cover_raster, opacity = 0.7,
                 colors = myPal2) %>%
  addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal2, domain = land_cover_type$Land_Cover),
    values =land_cover_type$Land_Cover,
    title = "Land Cover"
  )
```


```{r echo=FALSE}
ndvi_uganda <- raster("outputs/raster/ndvi_uganda_lower_res.tif")

ndvi_df <- raster::extract(ndvi_uganda, wgs84_coords)%>%
 tibble(ndvi = .,
    Sample_code  = wgs84_coords$Sample_code)

pal_ndvi <- colorNumeric("viridis", domain = c(0, 1))
leaflet() %>%
  addTiles() %>% 
  addRasterImage(ndvi_uganda, opacity = 0.7) %>%
  addLegend(
    "bottomright",
    pal = pal_ndvi,
    values = c(0, 1),
  )
```

```{r}
precip_uganda_resampled <- raster("data/raster/rainfall/rainfall_global.tif")
plot(precip_uganda_resampled)

precip_df <- raster::extract(precip_uganda_resampled, wgs84_coords)%>%
 tibble(precip = .,
    Sample_code  = wgs84_coords$Sample_code)
pal_precip <- colorNumeric("Spectral", domain = c(0, 1))
leaflet() %>%
  addTiles() %>% 
  addRasterImage(precip_uganda_resampled, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_precip,
    values = c(0, 1),
  )
precip_uganda_resampled %>% str()
```



```{r}
wgs84_coords %>%
  left_join(land_cover_df, by = "Sample_code") %>%
  left_join(land_degradation_df, by = "Sample_code") %>%
  left_join(ndvi_df, by = "Sample_code")%>%
  left_join(precip_df, by = "Sample_code") -> wgs84_coords
```


```{r}
pal_land_degradation <- colorFactor(myPal1[unique(land_degradation_df$ID)], 
                   domain = wgs84_coords$land_degra)

leaflet(wgs84_coords) %>%
  addTiles() %>%
  addCircleMarkers(
    ~lon, ~lat,
    color = "black",
    fillColor = ~pal_land_degradation(land_degra),
    radius = 6,
    weight = 2,
    stroke = T,
    fillOpacity = 0.8,
    popup = ~land_degra,
    label = ~Sample_code,
    group = ~land_degra
  ) %>%
   addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal1, domain = land_degradation_type$land_degra),
    values = land_degradation_type$land_degra,
    title = "Land Degradation"
  )


pal_land_cover <- colorFactor(myPal2[unique(land_cover_df$ID)], 
                   domain = wgs84_coords$Land_Cover)
leaflet(wgs84_coords) %>%
  addTiles() %>%
  addCircleMarkers(
    ~lon, ~lat,
    color = "black",
    fillColor = ~pal_land_cover(Land_Cover),
    radius = 6,
    weight = 2,
    stroke = T,
    fillOpacity = 0.8,
    popup = ~Land_Cover,
    label = ~Sample_code,
    group = ~Land_Cover
  ) %>%
  addLegend(
    "bottomright",
    pal = pal_land_cover,
    values = ~Land_Cover,
    title = "Land Cover"
  )
 
```

# JSDM

```{r}
amf_family_rare <- read.table("outputs/abundance_tables/Robust_MiSeq_18S_AMf_20241120_125_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.rarefy.family.txt")

family_tree <- read.tree("outputs/phylo_tree/Robust_MiSeq_18S_AMf_20241120_125_samples_Eukaryome_Glomeromycota_family.tree")
```


```{r}
library(Hmsc)

nChains = 12
test.run =F
if (test.run){
  #with this option, the vignette runs fast but results are not reliable
  thin = 5
  samples = 10
  transient = 5*thin
  verbose = 5*thin
} else {
  #with this option, the vignette evaluates slow but it reproduces the results of the
  #.pdf version
  thin = 10
  samples = 500
  transient = 1000*thin
  verbose = 1000*thin
}
amf_family_rare %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames("family") %>%
  colnames() -> family_matrix_samples
wgs84_coords %>%
  as.data.frame() %>%
  mutate(Sample_code= str_c("ROB_", Sample_code, "_s")) %>%
  filter(Sample_code %in% family_matrix_samples) -> covariate_df

amf_family_rare %>%
  remove_rownames() %>%
  column_to_rownames("family") %>%
  filter(rowSums(across(starts_with("ROB")) != 0) > 3)%>%
  t() -> Y_family
colSums(Y_family) %>% sort


Y_family %>% 
  as.data.frame() %>%
  rownames_to_column("soil_code") %>%
  pivot_longer(-soil_code) %>%
  mutate(value = case_when(value == 0 ~ NA,
                             .default = value)) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  column_to_rownames("soil_code") -> Y_family_trucated

Y_family_trucated %>% 
  as.data.frame() %>%
  pivot_longer(everything()) %>%
  ggplot() +
  facet_wrap(~name, scale = "free") +
  geom_histogram(aes(value))

family_tree_filter <- keep.tip(phy= family_tree, tip = colnames(Y_family_trucated))
dim(Y_family_trucated)
dim(covariate_df)
str(family_tree_filter)
colnames(Y_family_trucated)

studyDesign = data.frame(sample = rownames(Y_family_trucated), stringsAsFactors=TRUE)
rL = HmscRandomLevel(units = studyDesign$sample)
covariate_df %>%
  select(-geometry) %>%
  mutate(across(where(is.character), ~as.factor(.x))) -> covariate_df

covariate_df %>%
 pull(land_degra)  %>%
  table()
covariate_df %>%
 pull(Land_Cover)  %>%
  table()
mod_pois = Hmsc(Y = Y_family_trucated, XData = covariate_df, 
          XFormula = ~Land_Cover + land_degra + ndvi, 
          phyloTree =family_tree_filter,
          studyDesign = studyDesign,
          ranLevels = list(sample = rL),
          distr = "poisson")

# m_converg2 = sampleMcmc(mod_pois, thin = thin, samples = samples, transient = transient,
#                 nChains = nChains, verbose = verbose, nParallel = 12)
# save(m_converg2, file = "outputs/hmsc_family.Rdata")
load(file = "outputs/hmsc_family.Rdata")

```

```{r}
Y_family_trucated %>% 
  as.data.frame() %>%
  rownames_to_column("soil_code") %>%
  pivot_longer(-soil_code) %>%
  mutate(value = case_when(value != 0 ~ 1,
                             is.na(value) ~ 0)) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  column_to_rownames("soil_code") ->  Y_family_bin

m_bin = Hmsc(Y = Y_family_bin, XData = covariate_df, 
          XFormula = ~ Land_Cover + land_degra + ndvi , 
          phyloTree = family_tree_filter,
          studyDesign = studyDesign,
          ranLevels = list(sample = rL),
          distr = "probit")

# m_converg2_bin = sampleMcmc(m_bin, thin = thin, samples = samples, transient = transient,
#                 nChains = nChains, verbose = verbose, nParallel = 12)
# save(m_converg2_bin, file = "outputs/hmsc_family_bin_ndvi.Rdata")
load(file = "outputs/hmsc_family_bin_ndvi.Rdata")
```

```{r}
mpost2 = convertToCodaObject(m_converg2)
gelman.diag(mpost2$Beta,multivariate=FALSE) -> gelman_diag_df2
gelman_diag_df2$psrf
postBeta2 = getPostEstimate(m_converg2, parName = "Beta") 

mpost_bin2 = convertToCodaObject(m_converg2_bin)
gelman.diag(mpost_bin2$Beta,multivariate=FALSE) -> gelman_diag_bin_df2
gelman_diag_bin_df2$psrf
postBeta_bin2 = getPostEstimate(m_converg2_bin, parName = "Beta") 

plotBeta(m_converg2, post = postBeta2, param = "Mean", plotTree = TRUE, supportLevel = 0.95, split=.5, spNamesNumbers = c(T,T))
plotBeta(m_converg2_bin, post = postBeta_bin2, param = "Mean", plotTree = TRUE, supportLevel = 0.95, split=.5, spNamesNumbers = c(T,T))
```
Estimated response of each family to evironnemental variable. First figure show the responce for abundances and the second for occurence
```{r}
library(nlme)
library(purrr)
models = list(abund = m_converg2,bin = m_converg2_bin)

summary(corPagel(0.5, family_tree_filter, fixed=F,form = ~names(y)))
compute.padgel.lambda<- function(models, i){
  
  m = models[["bin"]]
  postBeta = getPostEstimate(m, parName="Beta")
  phyloTree = m$phyloTree
  nc = dim(postBeta$mean)[1]
  results = matrix(NA,nrow=nrow(postBeta2$mean), ncol = 4) %>%
    as.data.frame() %>%
    rename(lambda = "V1", p_value = "V2", model = "V3", covariate = "V4") 
  control <- glsControl(opt = "optim", msVerbose = TRUE)
  for(j in 1:nc){
    y = postBeta$mean[j,]
    m1<-gls(y~1)
    m2<-gls(y~1,correlation=corPagel(0.5, phyloTree, fixed=F,form = ~names(y)),control = control)
    z1=summary(m2)
    z2=anova(m1,m2)
    lambda= z1$modelStruct[1]$corStruct
    pvalue=z2$`p-value`[2]
    results$p_value[j]=pvalue
    results$lambda[j]=lambda
    results$model[j]=i
    results$covariate[j]= colnames(m$X)[j]
  }
  return(results)
}
any(branching.times(family_tree_filter) == 0)
c("abund", "bin") %>%
  purrr::map(\(x) compute.padgel.lambda(models, x)) %>%
  list_rbind() -> lambda_df

lambda_df %>%
  knitr::kable()

lambda_df%>%
  filter(p_value<=0.05) %>%
  ggplot()+
  geom_point(aes(lambda,covariate, col = model), cex = 4)+
   theme_classic()+
    theme(
          axis.title.y = element_blank(),
          axis.text.x = element_text(colour = "black", size=13, face="italic", angle = 90, vjust =0, hjust =1),
          axis.text.y = element_text(colour = "black", size=13, face="italic"),
          axis.title = element_text(colour = "black", size=16),
          panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          legend.title = element_text(colour = "black", size=16,
                                      hjust =0.5),
  
          legend.text = element_text(colour = "black", size=13))
  
```
Lambda is a metric that characterize phylogenetic signal. Here we charachterise how responce is related to phylogeny. When Lambda is close to 1 closelly related species will have the same responce to environmental variable. When Lambda is close to 0 response are to environmental variable are random across the phylogeny. Here occurrences of families are strongly phylogenically structure by environmental variable. However no explanatory variable show a significant abundance phylogeny structure.
Some reference for this metric : 

https://doi.org/10.1111/j.2041-210X.2012.00196.x
https://doi.org/10.1111/1365-2745.13839

```{r}
VP2 = computeVariancePartitioning(m_converg2,na.ignore =T) 
VP_bin2 = computeVariancePartitioning(m_converg2_bin,na.ignore= T) 

VP2$vals %>%
  as.data.frame %>%
  rownames_to_column("var") %>%
  pivot_longer(-var) %>%
  ggplot()+
  geom_col(aes(x= name, y= value, fill =var)) +
  theme_classic()+
    theme(
          axis.title.y = element_blank(),
          axis.text.x = element_text(colour = "black", size=13, face="italic", angle = 90, vjust =0, hjust =1),
          axis.text.y = element_text(colour = "black", size=13, face="italic"),
          axis.title = element_text(colour = "black", size=16),
          panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          legend.title = element_text(colour = "black", size=16,
                                      hjust =0.5),
  
          legend.text = element_text(colour = "black", size=13))


VP_bin2$vals %>%
  as.data.frame %>%
  rownames_to_column("var") %>%
  pivot_longer(-var) %>%
  ggplot()+
  geom_col(aes(x= name, y= value, fill =var)) +
  theme_classic()+
    theme(
          axis.title.y = element_blank(),
          axis.text.x = element_text(colour = "black", size=13, face="italic", angle = 90, vjust =0, hjust =1),
          axis.text.y = element_text(colour = "black", size=13, face="italic"),
          axis.title = element_text(colour = "black", size=16),
          panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          legend.title = element_text(colour = "black", size=16,
                                      hjust =0.5),
  
          legend.text = element_text(colour = "black", size=13))
```

```{r}
gelman.diag(mpost2$Rho)$psrf
summary(mpost2$Rho)

gelman.diag(mpost_bin2$Rho)$psrf
summary(mpost_bin2$Rho)
```

```{r}
library(corrplot)



OmegaCor = computeAssociations(m_converg2) 
supportLevel = 0.95 
toPlot = ((OmegaCor[[1]]$support>supportLevel) + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean 
corrplot(toPlot, method = "color", col=colorRampPalette(c("blue","white","red"))(200), tl.cex=.6, tl.col="black", title=paste("random effect level:", m_converg2$rLNames[1]), mar=c(0,0,1,0),order = "hclust")

OmegaCor_bin = computeAssociations(m_converg2_bin) 
supportLevel = 0.95 
toPlot = ((OmegaCor_bin[[1]]$support>supportLevel) + (OmegaCor_bin[[1]]$support<(1-supportLevel))>0)*OmegaCor_bin[[1]]$mean 
corrplot(toPlot, method = "color", col=colorRampPalette(c("blue","white","red"))(200), tl.cex=.6, tl.col="black", title=paste("random effect level:", m_converg2$rLNames[1]), mar=c(0,0,1,0),order = "hclust")
```



Extract x value at each positions. Coffee plantation occur only at specific Land cover types thus I filter location (pixels) keep only those land cover types
```{r}
land_degradation_raster_df <- rasterToPoints(land_degradation_raster) %>% 
  as.data.frame() %>%
  mutate(land_degra = factor(land_degra,labels = levels(land_degradation_raster)[[1]]$land_degra))
  

land_cover_raster_df <- rasterToPoints(land_cover_raster) %>%
  as.data.frame() %>%
  mutate(Land_Cover = factor(Land_Cover,labels = levels(land_cover_raster)[[1]]$Land_Cover))

ndvi_raster_df <- rasterToPoints(ndvi_uganda) %>% 
  as.data.frame()%>%
  rename(ndvi = "ndvi_uganda_lower_res")

data_spatial_df <- land_cover_raster_df %>%
  left_join(land_degradation_raster_df, by = join_by("x", "y")) %>%
  left_join(ndvi_raster_df, by = join_by("x", "y")) %>%
  mutate(Land_Cover = case_when(!(Land_Cover %in% unique(covariate_df$Land_Cover)) ~ NA,
                               .default = Land_Cover),
         Land_Cover = as.factor(as.character(Land_Cover))) %>%
  na.omit
```


```{r}
library(viridis)
# 
# data_spatial_df %>%
#   pivot_longer(c(Land_Cover,land_degra)) %>%
#   ggplot()+
#   facet_wrap(~name)+
#   geom_raster(aes(x,y,fill = value))
#  
#  data_spatial_df%>%
#    dplyr::select(x,y) %>%
#    as.matrix -> xy.grid
# 
#  data_spatial_df%>%
#    dplyr::select(-c(x,y)) -> XData.grid
# 
#  Gradient <- prepareGradient(mod_pois, XDataNew = XData.grid, sDataNew = list(route = xy.grid))
#  predY <- stats::predict(m_converg2, Gradient = Gradient)
# 
#  Gradient2 <- prepareGradient(m_bin, XDataNew = XData.grid, sDataNew = list(route = xy.grid))
#  predY2 <- stats::predict(m_converg2_bin, Gradient = Gradient2)
# 
#  EpredY <- predY[[1]]
#  EpredY2 <- predY2[[1]]
#  for(i in 2 :length(predY)){
#    EpredY <- EpredY + predY[[i]]
#    if(i%%1000 == 0){
#      print(i)
#    }
#  }
#  EpredY <- EpredY/length(predY)
# 
#  for(i in 2 :length(predY2)){
#    EpredY2 <- EpredY2 + predY2[[i]]
#    if(i%%1000 == 0){
#      print(i)
#    }
#  }
#  EpredY2 <- EpredY2/length(predY)
#  save(EpredY, file = "outputs/prediction_abudance.Rdata")
#  save(EpredY2, file = "outputs/prediction_occurence.Rdata")

load(file = "outputs/prediction_abudance.Rdata")
load(file = "outputs/prediction_occurence.Rdata")

data_spatial_df %>%
  bind_cols(EpredY) %>%
  pivot_longer(-c(x,y, Land_Cover, land_degra, ndvi)) %>%
  mutate(value = log10(value+1)) %>%
  ggplot()+
  facet_wrap(~name,)+
  geom_raster(aes(x,y,fill = value))+
  scale_fill_viridis(option = "cividis",direction = -1) +
  labs(x = " Lon", y = "Lat", fill = "Abundance (log10)") +
  theme_classic()+
  theme(aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        axis.title.y = element_text(colour = "black", size=14, face="italic"))

data_spatial_df %>%
  bind_cols(EpredY2) %>%
  pivot_longer(-c(x,y, Land_Cover, land_degra, ndvi)) %>%
  ggplot()+
  facet_wrap(~name)+
  geom_raster(aes(x,y,fill = value))+
  scale_fill_viridis(option = "magma",direction = -1) +
  labs(x = " Lon", y = "Lat", fill = "Occurence probability") +
  theme_classic()+
  theme(aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        axis.title.y = element_text(colour = "black", size=14, face="italic"))

data_spatial_df %>%
  bind_cols(EpredY2) %>%
  pivot_longer(-c(x,y, Land_Cover, land_degra)) %>%
  group_by(x,y) %>%
  summarise(richness = sum(value)) %>%
  ggplot()+
  geom_raster(aes(x,y,fill = richness))+
  scale_fill_viridis(option = "viridis",direction = -1) +
  labs(x = " Lon", y = "Lat", fill = "Family richness") +
  theme_classic()+
  theme(aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        axis.title.y = element_text(colour = "black", size=14, face="italic"))
```



