---
title: "jsdm_family"
output: html_document
date: "2025-05-27"
editor_options: 
  chunk_output_type: inline
---
```{r include=FALSE}
library(tidyverse)
library(raster)
library(mapview)
library(leaflet)
library(leafem)
library(rasterVis)
library(readxl)
library(sf)
library(pals)

library(Hmsc)
library(ape)

```

# Spatial covariate

## Sampling points
```{r echo=FALSE}
metadata <- read_xlsx(path = "data/Metadata_Doreen_070525.xlsx")

utm_coords <- metadata %>%
  mutate(
    zone = as.numeric(substr(easting, 1, 2)),
    band = substr(easting, 3, 3),
    easting = as.numeric(sub(".* ", "", easting)),
    northing = as.numeric(northing),

    hemisphere = ifelse(band == "N", "north", "south") # M = South Hemisphere
  )

# Function to convert a single row to lat/lon
convert_to_wgs84 <- function(zone, hemisphere, easting, northing,Sample_code) {
  epsg <- if (hemisphere == "north") {
    32600 + zone  # Northern hemisphere UTM EPSG codes
  } else {
    32700 + zone  # Southern hemisphere UTM EPSG codes
  }

  point <- st_sfc(st_point(c(easting, northing)), crs = epsg)
  point_wgs84 <- st_transform(point, crs = 4326)
  coords <- st_coordinates(point_wgs84)
  tibble( lon = coords[1], lat = coords[2],Sample_code = Sample_code)
}
utm_coords %>%
  mutate(zone = case_when(zone==32~ 36,
                             .default = zone),
          northing = case_when(Sample_code == "N17" ~ 86562,
                             .default = northing)) -> utm_coords
# Apply to each row
wgs84_coords <- pmap_dfr(utm_coords[, c("zone", "hemisphere", "easting", "northing","Sample_code")], convert_to_wgs84) %>%
  left_join(metadata, by = "Sample_code") %>%
  st_as_sf(coords = c("lon", "lat"),crs = 4326) %>%
  mutate(
     lon= st_coordinates(.)[,1],
     lat= st_coordinates(.)[,2]
  )


leaflet() %>% 
  addTiles() %>%
  addMarkers(data=wgs84_coords,~lon, ~lat, popup = ~paste(round(lat,2), round(lon,2)) , label = ~Sample_code)%>%
  addCircleMarkers(data = wgs84_coords, lng = ~lon,
                  lat = ~lat, radius = 8, color = "blue", fillOpacity = 0.8)
```

## Land degradation

This metric come from : https://gmesgeoportal.rcmrd.org/maps/8005edb1d35249e89f52260971c11c8b/about
key parameters used include: soil erosivity, slope, rainfall intensity, population and land cover

```{r echo=FALSE}
land_degradation_raster <- raster("outputs/raster/land_degradation_reclassify.tif")
land_degradation_type <- levels(land_degradation_raster)[[1]]
land_degradation_type <- land_degradation_type %>%
  mutate(land_degra = factor(land_degra, levels = c("Low", "Medium", "High")))
land_degradation_df <- raster::extract(land_degradation_raster,  wgs84_coords)%>%
  tibble(ID = .,
  Sample_code  = wgs84_coords$Sample_code)%>%
  left_join(land_degradation_type,  by = "ID")

myPal1 <- rev(RColorBrewer::brewer.pal('PRGn', n=3))


leaflet() %>%
  addTiles() %>% 
  addRasterImage(land_degradation_raster, opacity = 0.7,
                 color = myPal1) %>%
   addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal1, domain = land_degradation_type$land_degra),
    values = land_degradation_type$land_degra,
    title = "Land Degradation"
  )
```

## Land cover

This land cover metric is not ideal. Most of the sampling point fall in *Subsistence Cropland* (cf. 4 map below), leading to not enough replication of some land cover class for the model. We will probably replace this raster by agro-ecological zone one

```{r echo=FALSE}
land_cover_raster <- raster("outputs/raster/land_cover_lower_res.tif")

land_cover_type <- levels(land_cover_raster)[[1]] %>%
  dplyr::select(ID,Land_Cover) %>%
  mutate(Land_Cover = factor(Land_Cover, levels = levels(land_cover_raster)[[1]]$Land_Cover))


land_cover_df <- raster::extract(land_cover_raster, wgs84_coords) %>%
 tibble(ID = .,
    Sample_code  = wgs84_coords$Sample_code) %>%
    left_join(land_cover_type, by = "ID")

myPal2 <- pals::stepped2(n=18)

pal_land_cover <- colorFactor(myPal2[unique(land_cover_df$ID)], 
                   domain = wgs84_coords$Land_Cover)
leaflet() %>%
  addTiles() %>% 
  addRasterImage(land_cover_raster, opacity = 0.7,
                 colors = myPal2) %>%
  addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal2, domain = land_cover_type$Land_Cover),
    values =land_cover_type$Land_Cover,
    title = "Land Cover"
  )
```

## NDVI

Indice for vegetation quantity 


https://planetarycomputer.microsoft.com/dataset/sentinel-3-synergy-v10-l2-netcdf

```{r echo=FALSE}
ndvi_uganda <- raster("outputs/raster/ndvi_uganda_lower_res.tif")

ndvi_df <- raster::extract(ndvi_uganda, wgs84_coords)%>%
 tibble(ndvi = .,
    Sample_code  = wgs84_coords$Sample_code)

pal_ndvi <- colorNumeric("viridis", domain = c(0, 1))
leaflet() %>%
  addTiles() %>% 
  addRasterImage(ndvi_uganda, opacity = 0.7) %>%
  addLegend(
    "bottomright",
    pal = pal_ndvi,
    values = c(0, 1),
  )
```

## Precipitation

This is still work in progess, here the precipitation are for only **May 2023**. It is hard to get the data but I will try to obtain mean or sum precipitation for a hole year before June 2023.

```{r echo=FALSE}
precip_uganda_resampled <- raster("outputs/raster/precip_2023_lower_res.tif")

precip_df <- raster::extract(precip_uganda_resampled, wgs84_coords)%>%
 tibble(precip = .,
    Sample_code  = wgs84_coords$Sample_code)
pal_precip <- colorNumeric("Spectral", domain = c(0, 1))
leaflet() %>%
  addTiles() %>%
  addRasterImage(precip_uganda_resampled, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_precip,
    values = c(0, 1),
  )

```



```{r echo=FALSE}
wgs84_coords %>%
  left_join(land_cover_df, by = "Sample_code") %>%
  left_join(land_degradation_df, by = "Sample_code") %>%
  left_join(ndvi_df, by = "Sample_code") %>%
  left_join(precip_df, by = "Sample_code") %>%
  mutate(Land_Cover = as.factor(as.character(Land_Cover)),
    Land_Cover = factor(Land_Cover, levels = c("Dense Natural Forest", "Moderate Natural Forest", "Moderate Woodland",
                                                    "Perennial Commercial Cropland", "Subsistence Cropland" )))-> wgs84_coords
```

## Some value at the sample level (extract from the raster)

```{r echo=FALSE}
pal_land_degradation <- colorFactor(myPal1[unique(land_degradation_df$ID)], 
                   domain = wgs84_coords$land_degra)

leaflet(wgs84_coords) %>%
  addTiles() %>%
  addRasterImage(land_degradation_raster, opacity = 0.7,
                 color = myPal1) %>%
  addCircleMarkers(
    ~lon, ~lat,
    color = "black",
    fillColor = ~pal_land_degradation(land_degra),
    radius = 6,
    weight = 2,
    stroke = T,
    fillOpacity = 0.8,
    popup = ~land_degra,
    label = ~Sample_code,
    group = ~land_degra
  ) %>%
   addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal1, domain = land_degradation_type$land_degra),
    values = land_degradation_type$land_degra,
    title = "Land Degradation"
  )

pal_land_cover <- colorFactor(myPal2[sort(unique(land_cover_df$ID))], 
                   domain = unique(wgs84_coords$Land_Cover))
leaflet(wgs84_coords) %>%
    addTiles() %>%
  addRasterImage(land_cover_raster, opacity = 0.7,
                 color = myPal2) %>%

  addCircleMarkers(
    ~lon, ~lat,
    color = "black",
    fillColor = ~pal_land_cover(Land_Cover),
    radius = 6,
    weight = 2,
    stroke = T,
    fillOpacity = 0.8,
    popup = ~Land_Cover,
    label = ~Sample_code,
    group = ~Land_Cover
  ) %>%
  addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal2, domain = land_cover_type$Land_Cover),
    values =land_cover_type$Land_Cover,
    title = "Land Cover"
  )

```

# Joint species distribution models at the family level

We are looking for relationship between environmental variables from raster presented above and AMF OTU cluster at the family level. To do that we are using values of each 4 raster at the location of sampling points then I'm computing joint species distribution models with `Land cover + land degradation + NDVI + precipitation` as explaining variable. 
When the model is compute I can extrapolate spatial abundance and occurrence of each family according to raster values.

```{r echo=FALSE}
amf_family_rare <- read.table("outputs/abundance_tables/Robust_MiSeq_18S_AMf_20241120_125_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.rarefy.family.txt")

family_tree <- read.tree("outputs/phylo_tree/Robust_MiSeq_18S_AMf_20241120_125_samples_Eukaryome_Glomeromycota_family.tree")
family_tree$tip.label
family_tree <- root(family_tree, "c__Archaeosporomycetes_4")
```

Her ewe have cluster OTU at family levels, when family is not know for an OTU cluster at the highest taxonomic level known. In the names : the first letter indicate the level of taxonomic clustering (f = family, o = order, c = class).
Value indicate abundances (rarefied) of each family.

Both histogram below show the abundance count across samples. The fist one with zero (when OTU family are absent in a sample) and the second one without. 
Doing one model is not possible because of the zeros because it doesn't respect distribution condition.
I will do 2 model one with the **truncated zeros data** modeling **only the abundance**.
The second model is **presence-absence** transform data (binary) modeling **only the occurrences**. This is call a hurdle model where we use two models to study 2 process (abundance and occurrences) from the same data.

```{r echo=FALSE}
library(Hmsc)

nChains = 12
test.run = F
if (test.run){
  #with this option, the vignette runs fast but results are not reliable
  thin = 5
  samples = 10
  transient = 5*thin
  verbose = 5*thin
} else {
  #with this option, the vignette evaluates slow but it reproduces the results of the
  #.pdf version
  thin = 10
  samples = 500
  transient = 1000*thin
  verbose = 1000*thin
}
amf_family_rare %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames("family") %>%
  colnames() -> family_matrix_samples
wgs84_coords %>%
  as.data.frame() %>%
  mutate(Sample_code= str_c("ROB_", Sample_code, "_s")) %>%
  filter(Sample_code %in% family_matrix_samples) -> covariate_df

amf_family_rare %>%
  remove_rownames() %>%
  column_to_rownames("family") %>%
  filter(rowSums(across(starts_with("ROB")) != 0) > 3)%>%
  t() -> Y_family
colSums(Y_family) %>% sort


Y_family %>% 
  as.data.frame() %>%
  rownames_to_column("soil_code") %>%
  pivot_longer(-soil_code) %>%
  mutate(value = case_when(value == 0 ~ NA,
                             .default = value)) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  column_to_rownames("soil_code") -> Y_family_trucated
Y_family %>% 
  as.data.frame() %>%
  pivot_longer(everything()) %>%
  ggplot() +
  facet_wrap(~name, scale = "free") +
  geom_histogram(aes(value)) +
  labs(x = "abundances")

Y_family_trucated %>% 
  as.data.frame() %>%
  pivot_longer(everything()) %>%
  ggplot() +
  facet_wrap(~name, scale = "free") +
  geom_histogram(aes(value)) +
  labs(x = "abundances without zeros")
  

family_tree_filter <- keep.tip(phy= family_tree, tip = colnames(Y_family_trucated))

dim(Y_family_trucated)
dim(covariate_df)

colnames(Y_family_trucated)

studyDesign = data.frame(sample = rownames(Y_family_trucated), stringsAsFactors=TRUE)
rL = HmscRandomLevel(units = studyDesign$sample)
covariate_df %>%
  dplyr::select(-geometry) %>%
  mutate(across(where(is.character), ~as.factor(.x)),
         Land_Cover = factor(Land_Cover, levels = c("Dense Natural Forest", "Moderate Natural Forest", "Moderate Woodland",
                                                    "Perennial Commercial Cropland", "Subsistence Cropland" )),
         land_degra = factor(land_degra, levels = c("Low", "Medium", "High"))) -> covariate_df

levels(covariate_df$land_degra)
```


```{r echo=FALSE}
mod_pois = Hmsc(Y = Y_family_trucated, XData = covariate_df, 
          XFormula = ~Land_Cover + land_degra + ndvi + precip, 
          phyloTree =family_tree_filter,
          studyDesign = studyDesign,
          ranLevels = list(sample = rL),
          distr = "poisson")

# m_converg2 = sampleMcmc(mod_pois, thin = thin, samples = samples, transient = transient,
#                 nChains = nChains, verbose = verbose, nParallel = 12)
# save(m_converg2, file = "outputs/hmsc_family.Rdata")
load(file = "outputs/hmsc_family.Rdata")

```

```{r echo=FALSE}
Y_family_trucated %>% 
  as.data.frame() %>%
  rownames_to_column("soil_code") %>%
  pivot_longer(-soil_code) %>%
  mutate(value = case_when(value != 0 ~ 1,
                             is.na(value) ~ 0)) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  column_to_rownames("soil_code") ->  Y_family_bin

m_bin = Hmsc(Y = Y_family_bin, XData = covariate_df, 
          XFormula = ~ Land_Cover + land_degra + ndvi + precip, 
          phyloTree = family_tree_filter,
          studyDesign = studyDesign,
          ranLevels = list(sample = rL),
          distr = "probit")

# m_converg2_bin = sampleMcmc(m_bin, thin = thin, samples = samples, transient = transient,
#                 nChains = nChains, verbose = verbose, nParallel = 12)
# save(m_converg2_bin, file = "outputs/hmsc_family_bin_ndvi.Rdata")
load(file = "outputs/hmsc_family_bin_ndvi.Rdata")
```

```{r include=FALSE}
mpost2 = convertToCodaObject(m_converg2)
gelman.diag(mpost2$Beta,multivariate=FALSE) -> gelman_diag_df2
gelman_diag_df2$psrf
postBeta2 = getPostEstimate(m_converg2, parName = "Beta") 

mpost_bin2 = convertToCodaObject(m_converg2_bin)
gelman.diag(mpost_bin2$Beta,multivariate=FALSE) -> gelman_diag_bin_df2
gelman_diag_bin_df2$psrf
postBeta_bin2 = getPostEstimate(m_converg2_bin, parName = "Beta")
```

## Models results

### Family response to each varaibles

Estimated response of each family to environmental variable with phylogenetic tree. 

Response of abundance :

```{r echo=FALSE}
plotBeta(m_converg2, post = postBeta2, param = "Mean", plotTree = TRUE, supportLevel = 0.95, split=.5, spNamesNumbers = c(T,T))
```

Response of occurrences :

```{r echo=FALSE}

plotBeta(m_converg2_bin, post = postBeta_bin2, param = "Mean", plotTree = TRUE, supportLevel = 0.95, split=.5, spNamesNumbers = c(T,T))
```

### Phylogenetic signal

Abundances : 

```{r echo=FALSE}
gelman.diag(mpost2$Rho)$psrf
summary(mpost2$Rho)
```

Occurences : 

```{r echo=FALSE}
gelman.diag(mpost_bin2$Rho)$psrf
summary(mpost_bin2$Rho)
```


```{r include=FALSE}
library(nlme)
library(purrr)
models = list(abund = m_converg2,bin = m_converg2_bin)

summary(corPagel(0.5, family_tree_filter, fixed=F,form = ~names(y)))
compute.padgel.lambda<- function(models, i){
  
  m = models[[i]]
  postBeta = getPostEstimate(m, parName="Beta")
  phyloTree = m$phyloTree
  nc = dim(postBeta$mean)[1]
  results = matrix(NA,nrow=nrow(postBeta2$mean), ncol = 4) %>%
    as.data.frame() %>%
    rename(lambda = "V1", p_value = "V2", model = "V3", covariate = "V4") 
  control <- glsControl(opt = "optim", msVerbose = TRUE)
  for(j in 1:nc){
    y = postBeta$mean[j,]
    m1 <-gls(y~1)
    m2 <- gls(y~1,correlation=corPagel(0.5, phyloTree, fixed=F,form = ~names(y)),control = control)
    z1 = summary(m2)
    z2 = anova(m1,m2)
    lambda = z1$modelStruct[1]$corStruct
    pvalue = z2$`p-value`[2]
    results$p_value[j] = pvalue
    results$lambda[j] = lambda
    results$model[j] = i
    results$covariate[j] = colnames(m$X)[j]
  }
  return(results)
}
any(branching.times(family_tree_filter) == 0)
c("abund", "bin") %>%
  purrr::map(\(x) compute.padgel.lambda(models, x)) %>%
  list_rbind() -> lambda_df
```


```{r echo=FALSE}
lambda_df %>%
  knitr::kable()
```

Lambda is a metric that characterize phylogenetic signal. Here we characterize how response is related to phylogeny. When Lambda is close to 1 closely related species will have the same responses to environmental variable. When Lambda is close to 0 responses to environmental variables are random across the phylogeny. Here some reponses of occurrences and abundances are strongly phylogenically structure.

Some reference for this metric : 

https://doi.org/10.1111/j.2041-210X.2012.00196.x

https://doi.org/10.1111/1365-2745.13839

```{r echo=FALSE}
lambda_df%>%
  filter(p_value<=0.05) %>%
  ggplot()+
  geom_point(aes(lambda,covariate, col = model), cex = 4)+
   theme_classic()+
    theme(
          axis.title.y = element_blank(),
          axis.text.x = element_text(colour = "black", size=13, face="italic", angle = 90, vjust =0, hjust =1),
          axis.text.y = element_text(colour = "black", size=13, face="italic"),
          axis.title = element_text(colour = "black", size=16),
          panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          legend.title = element_text(colour = "black", size=16,
                                      hjust =0.5),
  
          legend.text = element_text(colour = "black", size=13))
  
```


### Variance partitonning

Variance explain by each variables for each family


abundances: 

```{r echo=FALSE, warning=FALSE}
VP2 = computeVariancePartitioning(m_converg2,na.ignore =T) 
VP_bin2 = computeVariancePartitioning(m_converg2_bin,na.ignore= T) 

VP2$vals %>%
  as.data.frame %>%
  rownames_to_column("var") %>%
  pivot_longer(-var) %>%
  ggplot()+
  geom_col(aes(x= name, y= value, fill =var)) +
  theme_classic()+
    theme(
          axis.title.y = element_blank(),
          axis.text.x = element_text(colour = "black", size=13, face="italic", angle = 90, vjust =0, hjust =1),
          axis.text.y = element_text(colour = "black", size=13, face="italic"),
          axis.title = element_text(colour = "black", size=16),
          panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          legend.title = element_text(colour = "black", size=16,
                                      hjust =0.5),
  
          legend.text = element_text(colour = "black", size=13))
```

Occurrences :

```{r echo=FALSE, warning=FALSE}
VP_bin2$vals %>%
  as.data.frame %>%
  rownames_to_column("var") %>%
  pivot_longer(-var) %>%
  ggplot()+
  geom_col(aes(x= name, y= value, fill =var)) +
  theme_classic()+
    theme(
          axis.title.y = element_blank(),
          axis.text.x = element_text(colour = "black", size=13, face="italic", angle = 90, vjust =0, hjust =1),
          axis.text.y = element_text(colour = "black", size=13, face="italic"),
          axis.title = element_text(colour = "black", size=16),
          panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
          legend.title = element_text(colour = "black", size=16,
                                      hjust =0.5),
  
          legend.text = element_text(colour = "black", size=13))
```


### Residuals correlation matrix

Correlation between each family when all the environmental variable have been accounted for.

Those figures show unexplained correlation caused by omitted biotic or abiotic variables

Abundances :

```{r echo=FALSE}
library(corrplot)



OmegaCor = computeAssociations(m_converg2) 
supportLevel = 0.95 
toPlot = ((OmegaCor[[1]]$support>supportLevel) + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean 
corrplot(toPlot, method = "color", col=colorRampPalette(c("blue","white","red"))(200), tl.cex=.6, tl.col="black", title=paste("random effect level:", m_converg2$rLNames[1]), mar=c(0,0,1,0),order = "hclust")
```

Occurrences : 

```{r echo=FALSE}
OmegaCor_bin = computeAssociations(m_converg2_bin) 
supportLevel = 0.95 
toPlot = ((OmegaCor_bin[[1]]$support>supportLevel) + (OmegaCor_bin[[1]]$support<(1-supportLevel))>0)*OmegaCor_bin[[1]]$mean 
corrplot(toPlot, method = "color", col=colorRampPalette(c("blue","white","red"))(200), tl.cex=.6, tl.col="black", title=paste("random effect level:", m_converg2$rLNames[1]), mar=c(0,0,1,0),order = "hclust")
```

### Spatial extrapolation

Coffee plantation occur only at specific Land cover types thus I filter location (pixels) keep only those land cover types (this is why we have big white area on land)

```{r echo=FALSE}
land_degradation_raster_df <- rasterToPoints(land_degradation_raster) %>% 
  as.data.frame() %>%
  mutate(land_degra = factor(land_degra,labels = levels(land_degradation_raster)[[1]]$land_degra))
  

land_cover_raster_df <- rasterToPoints(land_cover_raster) %>%
  as.data.frame() %>%
  mutate(Land_Cover = factor(Land_Cover,labels = levels(land_cover_raster)[[1]]$Land_Cover))

ndvi_raster_df <- rasterToPoints(ndvi_uganda) %>% 
  as.data.frame()%>%
  rename(ndvi = "ndvi_uganda_lower_res")

precip_raster_df <- rasterToPoints(precip_uganda_resampled) %>% 
  as.data.frame()%>%
  rename(precip = "precip_2023_lower_res")

data_spatial_df <- land_cover_raster_df %>%
  left_join(land_degradation_raster_df, by = join_by("x", "y")) %>%
  left_join(ndvi_raster_df, by = join_by("x", "y")) %>%
   left_join(precip_raster_df, by = join_by("x", "y")) %>%
  mutate(Land_Cover = case_when(!(Land_Cover %in% unique(covariate_df$Land_Cover)) ~ NA,
                               .default = Land_Cover),
         Land_Cover = as.factor(as.character(Land_Cover))) %>%
  na.omit
```


```{r echo=FALSE}
library(viridis)

# data_spatial_df %>%
#   pivot_longer(c(Land_Cover,land_degra)) %>%
#   ggplot()+
#   facet_wrap(~name)+
#   geom_raster(aes(x,y,fill = value))
# 
#  data_spatial_df%>%
#    dplyr::select(x,y) %>%
#    as.matrix -> xy.grid
# 
#  data_spatial_df%>%
#    dplyr::select(-c(x,y)) -> XData.grid
# 
#  Gradient <- prepareGradient(mod_pois, XDataNew = XData.grid, sDataNew = list(route = xy.grid))
#  predY <- stats::predict(m_converg2, Gradient = Gradient)
# 
#  Gradient2 <- prepareGradient(m_bin, XDataNew = XData.grid, sDataNew = list(route = xy.grid))
#  predY2 <- stats::predict(m_converg2_bin, Gradient = Gradient2)
# 
#  EpredY <- predY[[1]]
#  EpredY2 <- predY2[[1]]
#  for(i in 2 :length(predY)){
#    EpredY <- EpredY + predY[[i]]
#    if(i%%1000 == 0){
#      print(i)
#    }
#  }
#  EpredY <- EpredY/length(predY)
# 
#  for(i in 2 :length(predY2)){
#    EpredY2 <- EpredY2 + predY2[[i]]
#    if(i%%1000 == 0){
#      print(i)
#    }
#  }
# 
# EpredY2 <- EpredY2/length(predY)
# save(EpredY, file = "outputs/prediction_abudance.Rdata")
# save(EpredY2, file = "outputs/prediction_occurence.Rdata")

load(file = "outputs/prediction_abudance.Rdata")
load(file = "outputs/prediction_occurence.Rdata")

data_spatial_df %>%
  bind_cols(EpredY) %>%
  pivot_longer(-c(x,y, Land_Cover, land_degra, ndvi, precip)) %>%
  mutate(value = log10(value+1)) %>%
  ggplot()+
  facet_wrap(~name,)+
  geom_raster(aes(x,y,fill = value))+
  scale_fill_viridis(option = "cividis",direction = -1) +
  labs(x = " Lon", y = "Lat", fill = "Abundance (log10)") +
  theme_classic()+
  theme(aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        axis.title.y = element_text(colour = "black", size=14, face="italic"))

data_spatial_df %>%
  bind_cols(EpredY2) %>%
  pivot_longer(-c(x,y, Land_Cover, land_degra, ndvi, precip)) %>%
  ggplot()+
  facet_wrap(~name)+
  geom_raster(aes(x,y,fill = value))+
  scale_fill_viridis(option = "magma",direction = -1) +
  labs(x = " Lon", y = "Lat", fill = "Occurence probability") +
  theme_classic()+
  theme(aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        axis.title.y = element_text(colour = "black", size=14, face="italic"))

data_spatial_df %>%
  bind_cols(EpredY2) %>%
  pivot_longer(-c(x,y, Land_Cover, land_degra, ndvi, precip)) %>%
  group_by(x,y) %>%
  summarise(richness = sum(value)) %>%
  ggplot()+
  geom_raster(aes(x,y,fill = richness))+
  scale_fill_viridis(option = "viridis",direction = -1) +
  labs(x = " Lon", y = "Lat", fill = "Family richness") +
  theme_classic()+
  theme(aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        axis.title.y = element_text(colour = "black", size=14, face="italic"))
```



