---
title: "Untitled"
output: html_document
date: "2025-09-04"
---

```{r include=FALSE}
library(tidyverse)
library(raster)
library(mapview)
library(leaflet)
library(leafem)
library(rasterVis)
library(readxl)
library(sf)
library(pals)

library(Hmsc)
library(ape)

main_theme = theme_minimal() +
 theme(aspect.ratio = 1,
        axis.text.x = element_text(colour = "black", size=13, face="italic",angle =45, vjust =1, hjust =1),
        axis.text.y = element_text(colour = "black", size=13, face="italic"),
        axis.title = element_text(colour = "black", size=12),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        legend.title = element_text(colour = "black", size=12,
                                    hjust =0.5),

        legend.text = element_text(colour = "black", size=13))
```
# Spatial covariate

## Sampling points
```{r echo=FALSE}
metadata <- read_xlsx(path = "data/Metadata_Doreen_070525.xlsx")

utm_coords <- metadata %>%
  mutate(
    zone = as.numeric(substr(easting, 1, 2)),
    band = substr(easting, 3, 3),
    easting = as.numeric(sub(".* ", "", easting)),
    northing = as.numeric(northing),

    hemisphere = ifelse(band == "N", "north", "south") # M = South Hemisphere
  )

# Function to convert a single row to lat/lon
convert_to_wgs84 <- function(zone, hemisphere, easting, northing,Sample_code) {
  epsg <- if (hemisphere == "north") {
    32600 + zone  # Northern hemisphere UTM EPSG codes
  } else {
    32700 + zone  # Southern hemisphere UTM EPSG codes
  }

  point <- st_sfc(st_point(c(easting, northing)), crs = epsg)
  point_wgs84 <- st_transform(point, crs = 4326)
  coords <- st_coordinates(point_wgs84)
  tibble( lon = coords[1], lat = coords[2],Sample_code = Sample_code)
}
utm_coords %>%
  mutate(zone = case_when(zone==32~ 36,
                             .default = zone),
          northing = case_when(Sample_code == "N17" ~ 86562,
                             .default = northing)) -> utm_coords
# Apply to each row
wgs84_coords <- pmap_dfr(utm_coords[, c("zone", "hemisphere", "easting", "northing","Sample_code")], convert_to_wgs84) %>%
  left_join(metadata, by = "Sample_code") %>%
  st_as_sf(coords = c("lon", "lat"),crs = 4326) %>%
  mutate(
     lon= st_coordinates(.)[,1],
     lat= st_coordinates(.)[,2]
  )


leaflet() %>% 
  addTiles() %>%
  addMarkers(data=wgs84_coords,~lon, ~lat, popup = ~paste(round(lat,2), round(lon,2)) , label = ~Sample_code)%>%
  addCircleMarkers(data = wgs84_coords, lng = ~lon,
                  lat = ~lat, radius = 8, color = "blue", fillOpacity = 0.8)
```

## Land degradation

This metric come from : https://gmesgeoportal.rcmrd.org/maps/8005edb1d35249e89f52260971c11c8b/about
key parameters used include: soil erosivity, slope, rainfall intensity, population and land cover

```{r echo=FALSE}
land_degradation_raster <- raster("outputs/raster/land_degradation_reclassify.tif")
land_degradation_type <- levels(land_degradation_raster)[[1]]
land_degradation_type <- land_degradation_type %>%
  mutate(land_degra = factor(land_degra, levels = c("Low", "Medium", "High")))
land_degradation_df <- raster::extract(land_degradation_raster,  wgs84_coords)%>%
  tibble(ID = .,
  Sample_code  = wgs84_coords$Sample_code)%>%
  left_join(land_degradation_type,  by = "ID")

myPal1 <- rev(RColorBrewer::brewer.pal('PRGn', n=3))


leaflet() %>%
  addTiles() %>% 
  addRasterImage(land_degradation_raster, opacity = 0.7,
                 color = myPal1) %>%
   addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal1, domain = land_degradation_type$land_degra),
    values = land_degradation_type$land_degra,
    title = "Land Degradation"
  )
```

## Land cover

This land cover metric is not ideal. Most of the sampling point fall in *Subsistence Cropland* (cf. 4 map below), leading to not enough replication of some land cover class for the model. We will probably replace this raster by agro-ecological zone one

```{r echo=FALSE}
land_cover_raster <- raster("outputs/raster/land_cover_lower_res.tif")

land_cover_type <- levels(land_cover_raster)[[1]] %>%
  dplyr::select(ID,Land_Cover) %>%
  mutate(Land_Cover = factor(Land_Cover, levels = levels(land_cover_raster)[[1]]$Land_Cover))


land_cover_df <- raster::extract(land_cover_raster, wgs84_coords) %>%
 tibble(ID = .,
    Sample_code  = wgs84_coords$Sample_code) %>%
    left_join(land_cover_type, by = "ID")

myPal2 <- pals::stepped2(n=18)

pal_land_cover <- colorFactor(myPal2[unique(land_cover_df$ID)], 
                   domain = wgs84_coords$Land_Cover)
leaflet() %>%
  addTiles() %>% 
  addRasterImage(land_cover_raster, opacity = 0.7,
                 colors = myPal2) %>%
  addLegend(
    "bottomright",
    pal = colorFactor(palette=myPal2, domain = land_cover_type$Land_Cover),
    values =land_cover_type$Land_Cover,
    title = "Land Cover"
  )
```
## Cluster agroecologique

```{r}
cluster_agro_raster <- raster("outputs/raster/cluster_agro_lower_res.tif")
cluster_agro_type <- levels(cluster_agro_raster)[[1]] %>%
  dplyr::select(ID, Cluster) %>%
  mutate(Cluster = factor(Cluster, levels = levels(cluster_agro_raster)[[1]]$Cluster))

cluster_agro_df <- raster::extract(cluster_agro_raster, wgs84_coords) %>%
  tibble(ID = ., Sample_code = wgs84_coords$Sample_code) %>%
  left_join(cluster_agro_type, by = "ID")

myPal3 <- ggsci::pal_simpsons()(8)

leaflet() %>%
  addTiles() %>%
  addRasterImage(cluster_agro_raster, opacity = 0.7, colors = myPal3) %>%
  addLegend(
    "bottomright",
    pal = colorFactor(palette = myPal3, domain = cluster_agro_type$Cluster),
    values = cluster_agro_type$Cluster,
    title = "Land Cover"
  )
```


## NDVI

Indice for vegetation quantity 


https://planetarycomputer.microsoft.com/dataset/sentinel-3-synergy-v10-l2-netcdf

```{r echo=FALSE}
ndvi_uganda <- raster("outputs/raster/ndvi_uganda_lower_res.tif")

ndvi_df <- raster::extract(ndvi_uganda, wgs84_coords)%>%
 tibble(ndvi = .,
    Sample_code  = wgs84_coords$Sample_code)

pal_ndvi <- colorNumeric("viridis", domain = c(0, 1))
leaflet() %>%
  addTiles() %>% 
  addRasterImage(ndvi_uganda, opacity = 0.7) %>%
  addLegend(
    "bottomright",
    pal = pal_ndvi,
    values = c(0, 1),
  )
```

## Precipitation and Temperature max min

This is still work in progess, here the precipitation are for only **May 2023**. It is hard to get the data but I will try to obtain mean or sum precipitation for a hole year before June 2023.

```{r echo=FALSE}
precip_uganda_resampled <- raster("outputs/raster/precip_2023_lower_res.tif")

precip_df <- raster::extract(precip_uganda_resampled, wgs84_coords)%>%
 tibble(precip = .,
    Sample_code  = wgs84_coords$Sample_code)
pal_precip <- colorNumeric("Spectral", domain = c(0, 1))
leaflet() %>%
  addTiles() %>%
  addRasterImage(precip_uganda_resampled, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_precip,
    values = c(0, 1),
  )

```
```{r echo=FALSE}
tmax_uganda_resampled <- raster("outputs/raster/tmax_2023_lower_res.tif")

tmax_df <- raster::extract(tmax_uganda_resampled, wgs84_coords)%>%
 tibble(tmax = .,
    Sample_code  = wgs84_coords$Sample_code)
pal_t <- colorNumeric("Spectral", domain = c(0, 34))
leaflet() %>%
  addTiles() %>%
  addRasterImage(tmax_uganda_resampled, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_t,
    values = c(0, 34),
  )

```

```{r echo=FALSE}
tmin_uganda_resampled <- raster("outputs/raster/tmin_2023_lower_res.tif")

tmin_df <- raster::extract(tmin_uganda_resampled, wgs84_coords)%>%
 tibble(tmin = .,
    Sample_code  = wgs84_coords$Sample_code)
leaflet() %>%
  addTiles() %>%
  addRasterImage(tmin_uganda_resampled, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_t,
    values = c(0, 34),
  )

```

```{r}
delta_t_uganda <- tmax_uganda_resampled -tmin_uganda_resampled

tdelta_df <- raster::extract(delta_t_uganda, wgs84_coords)%>%
 tibble(tdelta = .,
    Sample_code  = wgs84_coords$Sample_code)
tdelta_df %>% summary()
leaflet() %>%
  addTiles() %>%
  addRasterImage(delta_t_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_t,
    values = c(9, 12),
  )
```

```{r echo=FALSE}
wind_2023 <- raster("outputs/raster/wind_2023.tif")
pal_wind <- colorNumeric("Spectral", domain = c(0, 4))
wind_df <- raster::extract(wind_2023, wgs84_coords)%>%
 tibble(wind = .,
    Sample_code  = wgs84_coords$Sample_code)
leaflet() %>%
  addTiles() %>%
  addRasterImage(wind_2023, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_wind,
    values = c(0, 4),
  )

```

```{r echo=FALSE}
altitude_uganda <- raster("outputs/raster/altitude_uganda.tif")
pal_alt <- colorNumeric("Spectral", domain = c(0, 4500))
altidute_df <- raster::extract(altitude_uganda, wgs84_coords)%>%
 tibble(altitude = .,
    Sample_code  = wgs84_coords$Sample_code)
leaflet() %>%
  addTiles() %>%
  addRasterImage(altitude_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_alt,
    values = c(0, 4),
  )

```

```{r echo=FALSE}
ph_uganda <- raster("outputs/raster/pH.tif")
pal_ph <- colorNumeric("Spectral", domain = c(0, 75))

ph_df <- raster::extract(ph_uganda, wgs84_coords)%>%
 tibble(pH = .,
    Sample_code  = wgs84_coords$Sample_code)

leaflet() %>%
  addTiles() %>%
  addRasterImage(ph_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_ph,
    values = c(0, 75),
  )

```
```{r echo=FALSE}
sand_uganda <- raster("outputs/raster/sand.tif")
pal_sand <- colorNumeric("Spectral", domain = c(0, 550))

sand_df <- raster::extract(sand_uganda, wgs84_coords)%>%
 tibble(sand = .,
    Sample_code  = wgs84_coords$Sample_code)

leaflet() %>%
  addTiles() %>%
  addRasterImage(sand_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_sand,
    values = c(0, 550),
  )

```

```{r echo=FALSE}
nitrogen_uganda <- raster("outputs/raster/nitrogen.tif")
pal_nitrogen <- colorNumeric("Spectral", domain = c(0, 748))

nitrogen_df <- raster::extract(nitrogen_uganda, wgs84_coords)%>%
 tibble(nitrogen = .,
    Sample_code  = wgs84_coords$Sample_code)

leaflet() %>%
  addTiles() %>%
  addRasterImage(nitrogen_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_nitrogen,
    values = c(0, 748),
  )

```


```{r echo=FALSE}
OC_uganda <- raster("outputs/raster/organic_carbone.tif")
pal_OC <- colorNumeric("Spectral", domain = c(0, 1305))

OC_df <- raster::extract(OC_uganda, wgs84_coords)%>%
 tibble(OC = .,
    Sample_code  = wgs84_coords$Sample_code)

leaflet() %>%
  addTiles() %>%
  addRasterImage(OC_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_OC,
    values = c(0, 1305),
  )

```

```{r}
clay_uganda <- raster("outputs/raster/clay.tif")

pal_clay <- colorNumeric("Spectral", domain = c(0, 484))

clay_df <- raster::extract(clay_uganda, wgs84_coords)%>%
 tibble(clay = .,
    Sample_code  = wgs84_coords$Sample_code)

leaflet() %>%
  addTiles() %>%
  addRasterImage(clay_uganda, opacity = 0.5)%>%
    addLegend(
    "bottomright",
    pal = pal_clay,
    values = c(0, 484),
  )
```

## Agroecological zone

```{r echo=FALSE}
wgs84_coords %>%
  left_join(cluster_agro_df, by = "Sample_code") %>%
  left_join(land_degradation_df, by = "Sample_code") %>%
  left_join(ndvi_df, by = "Sample_code") %>%
  left_join(precip_df, by = "Sample_code") %>%
  left_join(tmax_df, by = "Sample_code") %>%
  left_join(tmin_df, by = "Sample_code") %>%
  left_join(tdelta_df, by ="Sample_code") %>%
  left_join(wind_df, by ="Sample_code") %>%
  left_join(altidute_df, by ="Sample_code") %>% 
  left_join(ph_df, by ="Sample_code") %>%
  left_join(sand_df, by ="Sample_code") %>%
  left_join(nitrogen_df, by ="Sample_code") %>%
  left_join(OC_df, by ="Sample_code") %>%
  left_join(clay_df, by ="Sample_code") -> wgs84_coords

```


# Joint species distribution models at the family level

```{r echo=FALSE}
amf_family_decont <- read.table("outputs/abundance_tables/Robust_MiSeq_18S_AMf_20241120_125_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.rarefy.family.txt", header = TRUE, stringsAsFactors = FALSE)
family_tree <- read.tree("outputs/phylo_tree/Robust_MiSeq_18S_AMf_20241120_125_samples_Eukaryome_Glomeromycota_family.tree")

# Inspect and root tree
family_tree$tip.label
plot(family_tree)
family_tree <- root(family_tree, "c__Archaeosporomycetes_4")

# Clean tip labels (example transformation)
family_rename <- family_tree$tip.label %>%
  str_remove("f__") %>%
  str_replace_all("[a-z]__", "Unknown ") %>%
  str_replace("_(\\d+)$", " (\\1 OTU)")
length(family_rename)
family_tree$tip.label <- family_rename
plot(family_tree)
str(family_tree)
```


```{r}
# Prepare Y matrix for families
amf_family_decont <- amf_family_decont %>%
  mutate(family = family %>%
           str_remove("f__") %>%
           str_replace_all("[a-z]__", "Unknown ") %>%
           str_replace("_(\\d+)$", " (\\1 OTU)")) %>%
  remove_rownames() %>%
  column_to_rownames("family")

Y_family <- amf_family_decont %>%
  filter(rowSums(across(starts_with("ROB")) != 0) > 3) %>%
  t()

# Prepare Y without zeros as a truncated version
Y_family_trucated <- Y_family %>%
  as.data.frame() %>%
  rownames_to_column("soil_code") %>%
  pivot_longer(-soil_code) %>%
  mutate(value = ifelse(value == 0, NA, value)) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  column_to_rownames("soil_code")

# Filter tree to the families present in Y
family_tree_filter <- keep.tip(phy = family_tree, tip = colnames(Y_family))
Y_family <- Y_family[, family_tree_filter$tip.label]
Y_family_trucated <- Y_family_trucated[, family_tree_filter$tip.label]

# Sample names from Y
family_matrix_samples <- rownames(Y_family)

```


## Covariates

```{r}
library(GGally)
library(FactoMineR)
covariate_df <- wgs84_coords %>%
  as.data.frame() %>%
  mutate(Sample_code = str_c("ROB_", Sample_code, "_s")) %>%
  filter(Sample_code %in% family_matrix_samples) %>%
  dplyr::select(Sample_code,ndvi, precip, tmax, tmin,tdelta, wind, altitude, clay,OC,nitrogen,sand,pH)

covariate_df %>%
  dplyr::filter(Sample_code %in% rownames(Y_family)) %>%
  dplyr::slice(match(rownames(Y_family), Sample_code)) %>%
  column_to_rownames("Sample_code") %>%
  mutate(across(where(is.numeric), ~vegan::decostand(.x, method = "standardize")[,1])) %>%
  rownames_to_column("Sample_code") %>%
  mutate(Sample_code = as.factor(Sample_code)) -> covariate_standa_df
# write.table(covariate_df, "outputs/covariate_table.txt")

covariate_standa_df %>%
  select(-Sample_code) %>%
  ggpairs(progress = F)

covariate_standa_df %>%
  select(-Sample_code) %>%
  PCA()

covariate_standa_df %>%
  select(ndvi, precip, tmax, tdelta, wind, OC, nitrogen,clay, pH, sand) %>%
  ggpairs(progress = F)

covariate_standa_df %>%
  select(ndvi, precip, tmax, tdelta, wind, OC, nitrogen,clay, pH, sand) %>%
  PCA()

covariate_standa_df %>%
  filter(sand <= -4)

covariate_standa_df %>%
  select(ndvi, precip, tmax, tdelta, wind, OC, nitrogen,clay, pH, sand) %>%
  ggpairs(progress = F)
covariate_standa_df %>%
  select(ndvi, precip, tmax, tdelta, wind) %>%
  ggpairs(progress = F)
```





```{r echo=FALSE}
library(Hmsc)

nChains = 12
test.run = F
if (test.run){
  #with this option, the vignette runs fast but results are not reliable
  thin = 5
  samples = 10
  transient = 5*thin
  verbose = 5*thin
} else {
  #with this option, the vignette evaluates slow but it reproduces the results of the
  #.pdf version
  thin = 10
  samples = 750
  transient = 1000*thin
  verbose = 1000*thin
}
# covariate_standa_df %>%
#   filter(Sample_code != "ROB_M19_s")  -> covariate_standa_filter_df
# Y_family %>%
#   as.data.frame() %>%
#   rownames_to_column("Sample_code") %>%
#   filter(Sample_code != "ROB_M19_s") %>%
#   column_to_rownames("Sample_code") %>%
#   as.matrix() -> Y_family_filter
studyDesign <- data.frame(sample = rownames(Y_family), stringsAsFactors = TRUE)
rL <- HmscRandomLevel(units = studyDesign$sample)


```


```{r echo=FALSE}

# Y_family_filter %>% 
#   rowSums %>% 
#   as.data.frame() %>%
#   rename(readscount = ".") %>%
#   rownames_to_column("Sample_code") -> readscount_df
# 
# covariate_standa_filter_df %>%
#   left_join(readscount_df, by = "Sample_code") %>%
#   mutate(Sample_code = as.factor(Sample_code)) -> covariate_standa_filter_df
```


```{r echo=FALSE}
mod_pois_null <- Hmsc(
  Y = Y_family,
  XData =  covariate_standa_df ,
  XFormula = ~ 1,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_ndvi <- Hmsc(
  Y = Y_family,
  XData = covariate_standa_df,
  XFormula = ~ ndvi,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)


mod_pois_precip <- Hmsc(
  Y = Y_family,
  XData = covariate_standa_df,
  XFormula = ~ precip,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_tmax <- Hmsc(
  Y = Y_family,
  XData =  covariate_standa_df,
  XFormula = ~ tmax,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_wind  <- Hmsc(
  Y = Y_family,
  XData = covariate_standa_df,
  XFormula = ~ wind,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_full <- Hmsc(
  Y = Y_family,
  XData = covariate_standa_df,
  XFormula = ~ ndvi + precip + tmax +tdelta + wind,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_poly <- Hmsc(
  Y = Y_family,
  XData = covariate_standa_df,
  XFormula = ~poly(ndvi,2) + poly(precip,2) + poly(tmax,2) + poly(tdelta,2) + poly(wind,2),
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_null_log <- Hmsc(
  Y = log1p(Y_family),
  XData =  covariate_standa_df ,
  XFormula = ~ 1,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_full_log <- Hmsc(
  Y = log1p(Y_family),
  XData = covariate_standa_df,
  XFormula = ~ ndvi + precip + tmax +tdelta + wind,
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)

mod_pois_poly_log <- Hmsc(
  Y = log1p(Y_family),
  XData = covariate_standa_df,
  XFormula = ~poly(ndvi,2) + poly(precip,2) + poly(tmax,2) + poly(tdelta,2) + poly(wind,2),
  #phyloTree = family_tree_filter,
  studyDesign = studyDesign,
  ranLevels = list(sample = rL),
  distr = "poisson"
)


model_list <- list(
  mod_pois_null      = mod_pois_null,
  mod_pois_ndvi      = mod_pois_ndvi,
  mod_pois_precip    = mod_pois_precip,
  mod_pois_tmax      = mod_pois_tmax,
  mod_pois_wind      = mod_pois_wind,
  mod_pois_full      = mod_pois_full,
  mod_pois_poly      = mod_pois_poly,
  mod_pois_null_log  = mod_pois_null_log,
  mod_pois_full_log  = mod_pois_full_log,
  mod_pois_poly_log  = mod_pois_poly_log
)
```


```{r echo=FALSE}
compute.hmsc <- function(hmsc_mod, nfold = 10, nChains = 12,   test.run = T){

  if (test.run){
    #with this option, the vignette runs fast but results are not reliable
    thin = 5
    samples = 10
    transient = 5*thin
    verbose = 5*thin
  } else {
    #with this option, the vignette evaluates slow but it reproduces the results of the
    #.pdf version
    thin = 10
    samples = 1000
    transient = 1000*thin
    verbose = 1000*thin
  }
  
  m_converg <- sampleMcmc(
    hmsc_mod,
    thin = thin, samples = samples, transient = transient,
    nChains = nChains, verbose = verbose, nParallel = nChains)
  
  set.seed(92828)
  partition = createPartition(m_converg, nfolds = nfold, column = "sample")
  
  expl_power = computePredictedValues(m_converg, thin = thin, nParallel = nChains)
  MF_expl = evaluateModelFit(hM = m_converg, predY = expl_power) 
  
  #Predictive power based on cross-validation  

  pred_power = computePredictedValues(m_converg, partition = partition, thin = thin, nParallel = nChains)
  MF_pred = evaluateModelFit(hM = m_converg, predY = pred_power)
  
  mPredY <- apply(abind::abind(pred_power, along = 3), c(1, 2), matrixStats::mean2)
  
  colnames(mPredY) <- colnames(m_converg$Y)
  rownames(mPredY) <- rownames(m_converg$Y)
  
  list(hmsc_mod = hmsc_mod, m_converg = m_converg , mPredY = mPredY, MF_expl = MF_expl, MF_pred = MF_pred)
  }
# save(m_converg, file = "outputs/hmsc_family_full.Rdata")
# load(file = "outputs/hmsc_family_full.Rdata")

```

```{r}
library(purrr)
1:length(model_list) %>%
  purrr::map(\(x) compute.hmsc(model_list[[x]], nfold = 5, nChains = 14,   test.run = F)) -> list_mods_crossvalidation

```

```{r}

extract.pred.power <- function(mod_crossvalidation, mod_ID){

  mod_crossvalidation[["mPredY"]] %>%
      as.data.frame() %>%
      rownames_to_column("Sample_code") %>%
      pivot_longer(-Sample_code, names_to = "family_names", values_to = "pred_abund") -> mPredY_long
  
  mod_crossvalidation[["m_converg"]]$Y %>%
    as.data.frame() %>%
    rownames_to_column("Sample_code") %>%
    pivot_longer(-Sample_code, names_to = "family_names", values_to = "true_abund") -> true_data_long
  
  mPredY_long %>%
    left_join(true_data_long, by = join_by(family_names,Sample_code)) %>%
    mutate (mod_ID = mod_ID) -> abund_df
  
  abund_df
}

1:length(model_list) %>%
  purrr::map(\(x) extract.pred.power(list_mods_crossvalidation[[x]],mod_ID = x)) %>%
  bind_rows() %>%
  mutate(pred_abund = case_when(mod_ID %in% c(8,9,10) ~ exp(pred_abund)-1,
                          .default = pred_abund),
         true_abund = case_when(mod_ID %in% c(8,9,10) ~ exp(true_abund)-1,
                          .default = true_abund))-> mods_crossvalidation_df
mods_crossvalidation_df %>%
  filter(mod_ID == 1) %>% head
mods_crossvalidation_df %>%
  filter(mod_ID == 2) %>% head
mods_crossvalidation_df %>%
  filter(mod_ID == 10) %>% head

1:length(model_list)%>%
  purrr::map(\(x) list_mods_crossvalidation[[x]][["MF_pred"]] %>%
               as.data.frame %>%
               mutate(family_names = colnames(Y_family),
                      mod_ID = x)) %>%
  bind_rows() %>%
  mutate(RMSE = case_when(mod_ID %in% c(8,9,10) ~ exp(RMSE)+1,
                        .default = RMSE)) %>%
  left_join( mods_crossvalidation_df %>%
    group_by(family_names,mod_ID) %>%
    summarise(median_abund = median(true_abund, na.rm = T),
              mean_abund = mean(true_abund, na.rm = T),
              sd_abund = sd(true_abund, na.rm = T),
              median_pred = median(pred_abund, na.rm = T),
              mean_pred = mean(pred_abund, na.rm = T),
              sd_pred = sd(pred_abund, na.rm = T)), by = join_by(family_names== family_names, mod_ID==mod_ID)) %>%
  mutate(ratio_RMSE_sd = (RMSE)/sd_abund) -> recap_pred_df

1:length(model_list) %>%
  purrr::map(\(x) list_mods_crossvalidation[[x]][["MF_expl"]] %>%
               as.data.frame %>%
               mutate(family_names = colnames(Y_family),
                      mod_ID = x)) %>%
  bind_rows() %>%
  mutate(RMSE = case_when(mod_ID %in% c(8,9,10) ~ exp(RMSE)+1,
                          .default = RMSE))-> recap_expl_df

```

```{r}
recap_expl_df %>%
  group_by(mod_ID) %>%
  summarise(mean_rmse = mean(RMSE)) %>%
  ggplot() +
  geom_point(aes(mod_ID, mean_rmse)) +
  scale_y_log10() +
  main_theme

recap_expl_df %>%
  group_by(mod_ID) %>%
  summarise(mean_rmse_expl = mean(RMSE)) %>%
  left_join(recap_pred_df %>%
    group_by(mod_ID) %>%
    summarise(mean_rmse_pred = mean(RMSE)), by = "mod_ID") %>%
  ggplot() +
  geom_point(aes(mod_ID,mean_rmse_pred/mean_rmse_expl, col = as.factor(mod_ID)))+

  scale_y_log10() +
  scale_x_log10() 

```

```{r}
library(viridis)
recap_pred_df %>%
  #filter(!(mod_ID %in% c(1,2,3,4))) %>%
  ggplot() +
  geom_point(aes(family_names, ratio_RMSE_sd, col = as.factor(mod_ID)),cex = 3, alpha =0.8) +
  scale_y_log10() +
  scale_color_viridis(discrete = T)+
  main_theme

recap_pred_df %>%
  #filter(!(mod_ID %in% c(1,2,3,4))) %>%
  ggplot() +
  geom_point(aes(family_names, RMSE, col = as.factor(mod_ID)), cex = 4) +
  scale_y_log10() +
  scale_color_viridis(discrete = T)+
  main_theme

recap_pred_df %>%
  #filter(!(mean_pred>5000 & mod_ID ==4 )) %>%
  ggplot() +
  geom_abline(slope=1) +
  geom_point(aes(mean_abund, mean_pred, col = family_names), cex =3)+
  facet_wrap(~mod_ID, scale = "free")+
  scale_x_log10() +
  scale_y_log10() +
  main_theme

recap_pred_df %>%
  group_by(mod_ID) %>%
  summarise(mean(RMSE))
```

```{r}
mods_crossvalidation_df %>%
  ggplot()+
  facet_wrap(~mod_ID, scale = "free")+
  geom_point(aes(true_abund,pred_abund)) +
  geom_smooth(aes(true_abund, pred_abund), method = "lm") +
  geom_abline(slope=1) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x= "Observed abundances", y = "Predicted abudances") +
  main_theme


mods_crossvalidation_df %>%

  ggplot()+

  geom_point(aes(true_abund,pred_abund), cex = 4) +
  facet_wrap(~family_names)+
    #facet_grid(col = vars(family_names), row = vars(mod_ID), scale  = "free") +
  geom_abline(slope=1) +
    scale_x_log10() +
  scale_y_log10() +
  labs(x= "Observed abundances", y = "Predicted abudances")+
  main_theme

mods_crossvalidation_df %>%
 filter(mod_ID == 9, family_names != "Unknown Gigasporales (6 OTU)") %>%
  ggplot()+

  geom_point(aes(true_abund,pred_abund, col = log(pred_abund/(true_abund+1))), cex = 4) +
  facet_wrap(~family_names)+
    #facet_grid(col = vars(family_names), row = vars(mod_ID), scale  = "free") +
  geom_abline(slope=1) +
  labs(x= "Observed abundances", y = "Predicted abudances")+
  scale_x_log10() +
  scale_y_log10() +
  scale_color_gradientn(colors = topo.colors(50)) +
  main_theme


mods_crossvalidation_df %>%
 filter(mod_ID %in% c(9,6),family_names != "Unknown Gigasporales (6 OTU)") %>%
  ggplot()+

  geom_point(aes(true_abund,pred_abund, col = as.factor(mod_ID)), cex = 4, alpha = 0.6) +
  facet_wrap(~family_names)+
    #facet_grid(col = vars(family_names), row = vars(mod_ID), scale  = "free") +
  geom_abline(slope=1) +
  labs(x= "Observed abundances", y = "Predicted abudances")+
  scale_x_log10() +
  scale_y_log10() +
  scale_color_viridis(discrete = T) +
  main_theme

mods_crossvalidation_df %>%
  filter(mod_ID %in% c(8,9),family_names != "Unknown Gigasporales (6 OTU)") %>%
  mutate(mod_ID = case_when(mod_ID == 8 ~"null_mod",
                            mod_ID == 9 ~"add_mod")) %>%
  pivot_wider(names_from = mod_ID, values_from = pred_abund) %>%
  ggplot()+
  geom_point(aes(null_mod,add_mod, col = log(add_mod/(true_abund+1))), cex = 4) +
  facet_wrap(~family_names)+
    #facet_grid(col = vars(family_names), row = vars(mod_ID), scale  = "free") +
  geom_abline(slope=1) +
  labs(x= "Predicted null", y = "Predicted additive")+
  scale_x_log10() +
  scale_y_log10() +
  scale_color_gradientn(colors = topo.colors(50))
  main_theme

```

```{r include=FALSE}
list_mods_crossvalidation[[9]]$m_converg -> m_converg_selected
postBeta2 <- getPostEstimate(m_converg_selected, parName = "Beta")
beta_results <- postBeta2$mean
mpost2 <- convertToCodaObject(m_converg_selected)
gelman.diag(mpost2$Beta, multivariate = FALSE) -> gelman_diag_df2
gelman_diag_df2$psrf

beta_results <- postBeta2$mean
rownames(beta_results) <- colnames(m_converg_selected$X)
beta_results <- beta_results * ((postBeta2$support > 0.9) + (postBeta2$support < (1 - 0.9)) > 0)
beta_results <- t(beta_results)[family_tree_filter$tip.label, colnames(m_converg_selected$X)]
#colnames(beta_results) <- c("Intercept","NDVI", "Rain", "tmin", "tdelta", "wind") #  "Cluster B", "Cluster C", "Cluster E", "LD Medium", "LD High", 

```
## Models results

### family response to each varaibles

Estimated response of each family to environmental variable with phylogenetic tree. 

Response of abundance :

```{r echo=FALSE}
library(ggtree)
p <- ggtree(family_tree_filter, layout = "roundrect") +
  geom_tiplab(align = TRUE) +
  theme_tree()

gheatmap(p, beta_results, offset = 1, width = 1, font.size = 5,
         colnames_angle = -60, hjust = 0) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "grey90",
                       limits = c(min(beta_results), max(beta_results))) +
  scale_y_continuous(expand = c(0, 5)) +
  labs(fill = "Parameters\nvalues")
```

### Phylogenetic signal

Abundances : 

```{r echo=FALSE}
gelman.diag(mpost2$Rho)$psrf
summary(mpost2$Rho)
```

```{r include=FALSE}
library(nlme)
library(purrr)
# ------------------------------------------------------------------
# Compute Pagel's lambda for each covariate via nlme gls + corPagel
# ------------------------------------------------------------------
models <- list(abund = m_converg) # , bin = m_converg_bin

compute.padgel.lambda <- function(models, i) {
  m <- models[[i]]
  postBeta <- getPostEstimate(m, parName = "Beta")
  phyloTree <- m$phyloTree
  nc <- dim(postBeta$mean)[1]

  results <- matrix(NA, nrow = nrow(postBeta$mean), ncol = 4) %>%
    as.data.frame() %>%
    rename(lambda = "V1", p_value = "V2", model = "V3", covariate = "V4")

  control <- glsControl(opt = "optim", msVerbose = TRUE)

  for (j in 1:nc) {
    y <- postBeta$mean[j, ]
    m1 <- gls(y ~ 1)
    m2 <- gls(y ~ 1, correlation = corPagel(0.5, phyloTree, fixed = FALSE, form = ~names(y)), control = control)
    z1 <- summary(m2)
    z2 <- anova(m1, m2)
    lambda <- z1$modelStruct[1]$corStruct
    pvalue <- z2$`p-value`[2]

    results$p_value[j] <- pvalue
    results$lambda[j] <- lambda
    results$model[j] <- i
    results$covariate[j] <- colnames(m$X)[j]
  }
  return(results)
}

any(branching.times(family_tree_filter) == 0)

c("abund") %>%#, "bin"
  purrr::map(\(x) compute.padgel.lambda(models, x)) %>%
  list_rbind() -> lambda_df


```


```{r echo=FALSE}
lambda_df %>%
  knitr::kable()
```

Lambda is a metric that characterize phylogenetic signal. Here we characterize how response is related to phylogeny. When Lambda is close to 1 closely related species will have the same responses to environmental variable. When Lambda is close to 0 responses to environmental variables are random across the phylogeny. Here some reponses of occurrences and abundances are strongly phylogenically structure.

Some reference for this metric : 

https://doi.org/10.1111/j.2041-210X.2012.00196.x

https://doi.org/10.1111/1365-2745.13839

```{r echo=FALSE}
lambda_df %>%
  filter(p_value <= 0.05) %>%
  ggplot() +
  geom_point(aes(lambda, covariate, col = model), cex = 4) +
  theme_classic() +
  theme(
    axis.title.y = element_blank(),
    axis.text.x = element_text(colour = "black", size = 13, face = "italic", angle = 90, vjust = 0, hjust = 1),
    axis.text.y = element_text(colour = "black", size = 13, face = "italic"),
    axis.title = element_text(colour = "black", size = 16),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    legend.title = element_text(colour = "black", size = 16, hjust = 0.5),
    legend.text = element_text(colour = "black", size = 13)
  )
```


### Variance partitonning

Variance explain by each variables for each family


abundances: 

```{r echo=FALSE, warning=FALSE}
VP2 <- computeVariancePartitioning(m_converg_selected, na.ignore = TRUE)


VP2$vals %>%
  as.data.frame() %>%
  rownames_to_column("var") %>%
  mutate(var = factor(var, levels = c("ndvi", "precip", "tmax", "tdelta", "wind", "Random: sample"))) %>%
  pivot_longer(-var) %>%
  ggplot() +
  geom_col(aes(x = name, y = value, fill = var)) +
  theme_classic() +
  theme(
    axis.title.y = element_blank(),
    axis.text.x = element_text(colour = "black", size = 9, face = "italic", angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_text(colour = "black", size = 13, face = "italic"),
    axis.title = element_text(colour = "black", size = 16),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    legend.title = element_text(colour = "black", size = 16, hjust = 0.5),
    legend.text = element_text(colour = "black", size = 13)
  ) +
  ggsci::scale_fill_simpsons()

```


### Residuals correlation matrix

Correlation between each family when all the environmental variable have been accounted for.

Those figures show unexplained correlation caused by omitted biotic or abiotic variables

Abundances :

```{r echo=FALSE}
library(corrplot)

OmegaCor <- computeAssociations(m_converg_selected)
supportLevel <- 0.95
toPlot <- ((OmegaCor[[1]]$support > supportLevel) + (OmegaCor[[1]]$support < (1 - supportLevel)) > 0) * OmegaCor[[1]]$mean

corrplot(toPlot, method = "color", col = colorRampPalette(c("blue", "white", "red"))(200),
         tl.cex = 0.6, tl.col = "black",
         title = paste("random effect level:", m_converg_selected$rLNames[1]),
         mar = c(0, 0, 1, 0), order = "hclust")
```

```{r echo=FALSE}

ndvi_raster_df <- rasterToPoints(ndvi_uganda) %>%
  as.data.frame() %>%
  rename(ndvi = "ndvi_uganda_lower_res")

precip_raster_df <- rasterToPoints(precip_uganda_resampled) %>%
  as.data.frame() %>%
  rename(precip = "precip_2023_lower_res")

tmax_raster_df <- rasterToPoints(tmax_uganda_resampled) %>%
  as.data.frame() %>%
  rename(tmax = "tmax_2023_lower_res")

tdelta_raster_df <- rasterToPoints(delta_t_uganda) %>%
  as.data.frame() %>%
  rename(tdelta = "layer")

wind_raster_df <- rasterToPoints(wind_2023) %>%
  as.data.frame() %>%
  rename(wind = "wind_2023")

data_spatial_df <- ndvi_raster_df %>%
  left_join(precip_raster_df, by = join_by("x", "y")) %>%
  left_join(tmax_raster_df, by = join_by("x", "y")) %>%
  left_join(tdelta_raster_df, by = join_by("x", "y")) %>%
  left_join(wind_raster_df, by = join_by("x", "y")) %>% 
  na.omit()%>%
  mutate(across(c(ndvi, precip, tmax, tdelta, wind), ~vegan::decostand(.x, method = "standardize")[,1]))
```

```{r}

data_spatial_df %>%
  pivot_longer(-c(x,y)) %>%
  ggplot() +
  facet_wrap(~name, scale = "free") +
  geom_raster(aes(x, y, fill = value))

# Prepare grid for gradient predictions
xy.grid <- data_spatial_df %>%
  dplyr::select(x, y) %>%
  as.matrix()

XData.grid <- data_spatial_df %>%
  dplyr::select(-c(x, y))

Gradient <-  prepareGradient(list_mods_crossvalidation[[9]]$hmsc_mod, XDataNew = XData.grid, sDataNew = list(route = xy.grid))
predY <- stats::predict(m_converg_selected, Gradient = Gradient)

EpredY_map <- predY[[1]]

str(predY)
# Average over predictions
if (length(predY) > 1) {
  for (i in 2:length(predY)) {
    EpredY_map <- EpredY_map + predY[[i]]
    if (i %% 1000 == 0) print(i)
  }
  EpredY_map <- EpredY_map / length(predY)
}
```

```{r}
library(viridis)
recap_df %>%
 filter(ratio_RMSE_sd<2) %>%
  pull(family_names) -> good_prediction
data_spatial_df %>%
    bind_cols(EpredY_map) %>%
    pivot_longer(-c(x, y, ndvi, precip, tmax, tdelta, wind)) %>%
    filter(name %in% good_prediction[1]) %>%
    mutate(value = value) %>%
    ggplot() +
    facet_wrap(~name) +
    geom_raster(aes(x, y, fill = value)) +
    scale_fill_viridis(option = "magma") +
    labs(x = "Lon", y = "Lat", fill = "Abundance (log10)") +
    theme_classic() +
    theme(
      aspect.ratio = 1,
      plot.margin = margin(0, 0, 0, 0),
      axis.title.x = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
      axis.title.y = element_text(colour = "black", size = 14, face = "italic")
    )
data_spatial_df %>%
    bind_cols(EpredY_map) %>%
    pivot_longer(-c(x, y, ndvi, precip, tmax, tdelta, wind)) %>%
    filter(name %in% good_prediction[2]) %>%
    mutate(value = value) %>%
    ggplot() +
    facet_wrap(~name) +
    geom_raster(aes(x, y, fill = value)) +
    scale_fill_viridis(option = "magma") +
    labs(x = "Lon", y = "Lat", fill = "Abundance (log10)") +
    theme_classic() +
    theme(
      aspect.ratio = 1,
      plot.margin = margin(0, 0, 0, 0),
      axis.title.x = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
      axis.title.y = element_text(colour = "black", size = 14, face = "italic")
    )
data_spatial_df %>%
    bind_cols(EpredY_map) %>%
    pivot_longer(-c(x, y, ndvi, precip, tmax, tdelta, wind)) %>%
    filter(name %in% good_prediction[3]) %>%
    mutate(value = value) %>%
    ggplot() +
    facet_wrap(~name) +
    geom_raster(aes(x, y, fill = value)) +
    scale_fill_viridis(option = "magma") +
    labs(x = "Lon", y = "Lat", fill = "Abundance (log10)") +
    theme_classic() +
    theme(
      aspect.ratio = 1,
      plot.margin = margin(0, 0, 0, 0),
      axis.title.x = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
      axis.title.y = element_text(colour = "black", size = 14, face = "italic")
    )
data_spatial_df %>%
    bind_cols(EpredY_map) %>%
    pivot_longer(-c(x, y, ndvi, precip, tmax, tdelta, wind)) %>%
    filter(name %in% good_prediction[4]) %>%
    mutate(value = value ) %>%
    ggplot() +
    facet_wrap(~name) +
    geom_raster(aes(x, y, fill = value)) +
    scale_fill_viridis(option = "magma") +
    labs(x = "Lon", y = "Lat", fill = "Abundance (log10)") +
    theme_classic() +
    theme(
      aspect.ratio = 1,
      plot.margin = margin(0, 0, 0, 0),
      axis.title.x = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
      axis.title.y = element_text(colour = "black", size = 14, face = "italic")
    )

```


